---
title: "EDA"
author: "Chenxu Wang"
#date: "today"
date: "`r Sys.Date()`"
# this style requires installing rmdformats package 
output:
  html_document:
    code_folding: hide
    number_sections: true
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---
```{r init, include=FALSE}
# some of common options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
library(ezids)
library(ggplot2)
library(ggpubr)
# knitr::opts_chunk$set(warning = F, results = "markup", message = F)
knitr::opts_chunk$set(warning = F, results = "hide", message = F, echo = T)
options(scientific=T, digits = 3) 
# options(scipen=9, digits = 3) 
# ‘scipen’: integer. A penalty to be applied when deciding to print numeric values in fixed or exponential notation.  Positive values bias towards fixed and negative towards scientific notation: fixed notation will be preferred unless it is more than ‘scipen’ digits wider.
# use scipen=999 to prevent scientific notation at all times
```

# Chapter 1: Introudction of cardiovascular disease

## 1.1 Basic Information

According to the American Heart Association, cardiovascular disease can refer to a number of conditions, including heart disease, heart attack, stroke, heart failure, arrhythmia, and heart valve problems. Heart disease and stroke are the leading causes of death worldwide, ranked first and second respectively. The World Heath Organization cites unhealthy diet, physical inactivity, tobacco and alcohol use as the most important behavioral risk factors of heart disease and stroke risk. As a result, patients with cardiovascular disease often present with raised blood pressure, raised blood glucose, raised blood lipids, and BMI in the overweight or obese range.

## 1.2 The purpose of project 

Based on a report published by Centers for Disease Control and Prevention, about 610,000 people die of heart disease in the United States every year: that is 1 in every 4 deaths. 
So, in this project we aim to answer below questions by exploring data: (The smart question)

# Chapter 2: Description of Dataset
## About Dataset
The Cardiovascular Disease dataset was obtained from Kaggle   (https://www.kaggle.com/datasets/sulianova/cardiovascular-disease-dataset). There are 70000 total values (observations) in 11 features, such as age, gender, systolic blood pressure, diastolic blood pressure, and etc. The target class “cardio” equals to 1, when patient has cardiovascular disease, and it’s 0, if patient is healthy.In this data, number of people people with cardiovascular disease are 34,979 and without are 35,021. 

There are 3 types of input features:

Objective: factual information
Examination: results of medical examination
Subjective: information given by the patient.



## Variables
```{r import, results='markup'}
cardio_orig<-read.csv("cardio_train.csv", header = TRUE, sep = ";")
str(cardio_orig)
```
* `id `: ID number
* `age`: Age (in days and we will transfer to years)
* `gender`: Women(1) or Men(2)
* `height`: Height(cm)
* `weight`: Weight(kg)
* `ap_hi`: Systolic blood pressure
* `ap_lo`: Astolic blood pressure
* `cholesterol`: Cholesterol(1: normal, 2: above normal, 3: well above normal)
* `gluc`: Glucose(1: normal, 2: above normal, 3: well above normal)
* `smoke`: whether patient smokes or not(0:NO,1:Yes)
* `alco`: Alcohol intake (0:NO,1:Yes)
* `active`: Physical activity (0:NO,1:Yes)
* `cardio`:  The target class “cardio” equals to 1, when patient has cardiovascular disease, and it’s 0, if patient is healthy.

## Remove outliers
```{r removeoutliers}
cardio_remove = outlierKD2(cardio_orig, ap_hi, TRUE, TRUE, TRUE, TRUE) 
cardio_orig = outlierKD2(cardio_remove, ap_lo, TRUE, TRUE, TRUE, TRUE)
```
With minimum as low as -150 and maximum as high as 16020
Blood pressure outside the range 0-400 means person is either dead or just about to die
We have removed records with these outliers from the data


 

```{r asfactor}

cardio_orig$age=cardio_orig$age/365 # transfer days to years

cardio <- transform(cardio_orig,gender=as.factor(gender),
                      alco = as.factor(alco),
                      smoke = as.factor(smoke), 
                      active = as.factor(active), 
                      cholesterol=as.factor(cholesterol), 
                      gluc= as.factor(gluc), 
                      cardio=as.factor(cardio))
str(cardio)
```

```{r cleanNA, include=FALSE}
na_cardio<- sum(is.na(cardio))
na_cardio # there are 0 NA in cardio
cardio <- na.omit(cardio)
sum(is.na(cardio))
```

```{r results=TRUE}
str(cardio)
```
For our exploratory data analysis, we did some preprocessing including cleaning up and converting.First of all, we remove the outliers in both blood pressure columns. Then we transfer days to years for age.We dropped “NA” values from the dataset to simplify our analysis; we converted some variables into factor variables except `id`, `age`, `height`, `weight`, `ap_hi`,`ap_lo`

# Chapter 3: Categorical Variables EDA  

A brief overview of the dataset tells that 

## Present disease vs Absent disease
### What are cardiovascular disease patient look like?



### Contrast present with absent

```{r, results='markup'}

gender_1 <- ggplot(cardio, aes(x=gender, fill=cardio)) +
                   geom_bar(position="fill") +
                   scale_fill_manual(values=c("pink3", "steelblue")) +
                   scale_x_discrete(labels=c("1" = "Women", "2" = "Man")) +
                   labs(title="gender", x="", y="Percentage") +
                   theme(legend.position="top")

cholesterol_1 <- ggplot(cardio, aes(x=cholesterol, fill= cardio)) +
                   geom_bar(position="fill") +
                   scale_fill_manual(values=c("pink3", "steelblue")) +
                   scale_x_discrete(labels=c("1" = "normal", "2" = "above normal","3" = "well above normal")) +
                   labs(title="cholesterol", x="", y="Percentage") +
                   theme(legend.position="top")
gluc_1<- ggplot(cardio, aes(x=gluc, fill= cardio)) +
                   geom_bar(position="fill") +
                   scale_fill_manual(values=c("pink3", "steelblue")) +
                   scale_x_discrete(labels=c("1" = "normal", "2" = "above normal","3" = "well above normal")) +
                   labs(title="gluc", x="", y="Percentage") +
                   theme(legend.position="top")

smoke_1<- ggplot(cardio, aes(x=smoke, fill= cardio)) +
                   geom_bar(position="fill") +
                   scale_fill_manual(values=c("pink3", "steelblue")) +
                   scale_x_discrete(labels=c("0" = "No smoking", "1" = "Smoking")) +
                   labs(title="smoke", x="", y="Percentage") +
                   theme(legend.position="top")

alco_1<- ggplot(cardio, aes(x=alco, fill= cardio)) +
                   geom_bar(position="fill") +
                   scale_fill_manual(values=c("pink3", "steelblue")) +
                   scale_x_discrete(labels=c("0" = "No alcohol intake", "1" = "Alcohol intake")) +
                   labs(title="alco", x="", y="Percentage") +
                   theme(legend.position="top")

active_1<- ggplot(cardio, aes(x=active, fill= cardio)) +
                   geom_bar(position="fill") +
                   scale_fill_manual(values=c("pink3", "steelblue")) +
                   scale_x_discrete(labels=c("0" = "No physical activity", "1" = "Physical activity" )) +
                   labs(title="active", x="", y="Percentage") +
                   theme(legend.position="top")

ggarrange(gender_1, cholesterol_1,gluc_1, smoke_1, alco_1,active_1)

```
According to the plot above, we can draw the conclusion that 
*It's close to half the female and male population have cardiovascular disease.
*The higher the degree of cholesterol and glucose ， more population have this disease.
*It is surprised that whether patient smokes or not and consumes alcohol or not, Almost half of them have this disease.
* Do physical exercise can slighlt help us decrease the probability of cardiovascular disease.

# Chapter 4: Continuous Variables EDA
##4.1 KDE plot of continuous variables
The following plots are stacked for better visualization of the distribution.
```{r,results='markup'}
age_kdeplot <- ggplot(cardio, aes(x = age, fill= cardio)) + 
  geom_density(position = "stack") +
  geom_vline(aes(xintercept=mean(age)), color="black", linetype="dashed", size=1) +
  xlab("Age ") + ylab("Percentage") + 
  ggtitle(expression(atop("Age Distribution (stacked)")))+
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_fill_manual(values=c("pink3", "steelblue"),labels=c("Absent", "Present")) + 
  xlim(20, 70)
age_kdeplot

```
Above plot is stacked for better visualization of the distribution.
The percentage of disease shows that the age of patients with disease mainly concentrate on around 50.


```{r,results='markup'}
height_kdeplot<-  ggplot(cardio, aes(x = height, fill= cardio)) + 
  geom_density(position = "stack") +
  geom_vline(aes(xintercept=mean(height)), color="black", linetype="dashed", size=1) +
  xlab("Height(in cm) ") + ylab("Proportion") + 
  ggtitle(expression(atop("Height Distribution (stacked)")))+
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_fill_manual(values=c("pink3", "steelblue"),labels=c("Absent", "Present")) + 
  xlim(10, 200)

height_kdeplot
```

  
```{r,results='markup'}
weight_kdeplot<- ggplot(cardio, aes(x = weight, fill= cardio)) + 
  geom_density(position = "stack") +
  geom_vline(aes(xintercept=mean(weight)), color="black", linetype="dashed", size=1) +
  xlab("Weight(in kg) ") + ylab("Proportion") + 
  ggtitle(expression(atop("Weight Distribution (stacked)")))+
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_fill_manual(values=c("pink3", "steelblue"),labels=c("Absent", "Present")) + 
  xlim(0, 200)

weight_kdeplot
```



```{r,results='markup'}
ap_hi_kdeplot<- ggplot(cardio, aes(x = ap_hi, fill= cardio)) + 
  geom_density(position = "stack") +
  geom_vline(aes(xintercept=mean(ap_hi)), color="black", linetype="dashed", size=1) +
  xlab("Systolic blood pressure(ap_hi) ") + ylab("Proportion") + 
  ggtitle(expression(atop("Systolic blood pressure Distribution (stacked)")))+
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_fill_manual(values=c("pink3", "steelblue"),labels=c("Absent", "Present")) + 
  xlim(50, 250)

ap_hi_kdeplot
```


```{r,results='markup'}
ap_lo_kdeplot<-ggplot(cardio, aes(x = ap_lo, fill= cardio)) + 
  geom_density(position = "stack") +
  geom_vline(aes(xintercept=mean(ap_lo)), color="black", linetype="dashed", size=1) +
  xlab("Diastolic blood pressure(ap_lo)) ") + ylab("Proportion") + 
  ggtitle(expression(atop("Diastolic blood pressure Distribution (stacked)")))+
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_fill_manual(values=c("pink3", "steelblue"),labels=c("Absent", "Present")) + 
  xlim(50, 150)

ap_lo_kdeplot
```


```{r,results='markup'}

ggarrange(age_kdeplot,height_kdeplot,weight_kdeplot,ap_lo_kdeplot,ap_hi_kdeplot)
```

## Classification Tree

This model is Classification Tree model. We could use the classification tree model to predict the absence and presence of disease visually. It is very convenient. And we will discuss the performance of this model below.

First step, cleaning the dataset for preparing to build the model. Ensure the variavles are all numerical variables. And delete the ID, which has no relationship to the results.

```{r , results='markup'}



#read in original data set
cardio_tree_orig = data.frame(read.csv("cardio_train.csv", header = TRUE, sep=';'))

#remove id column, convert age to years, add bmi col
cardio_tree_new = cardio_tree_orig[-c(1)]
cardio_tree_new$age = cardio_tree_new$age/365
cardio_tree_new$bmi = cardio_tree_new$weight/(cardio_tree_new$height/100)**2

#remove outliers from bmi, api_hi and api_lo
cardio_clean_1 = outlierKD2(cardio_tree_new, bmi, TRUE, FALSE, FALSE, FALSE) 
cardio_clean_2 = outlierKD2(cardio_clean_1, ap_hi, TRUE, FALSE, FALSE, FALSE) 
cardio_clean_3 = outlierKD2(cardio_clean_2, ap_lo, TRUE, FALSE, FALSE, FALSE)

#remove NA rows
cardio_clean_3 = na.omit(cardio_clean_3)
cardio_clean_tree_final <- transform(cardio_clean_3, cardio=as.factor(cardio))
cardio_tree = cardio_clean_tree_final[,c(1,2,3,4,13,5,6,7,8,9,10,11,12)]
str(cardio_tree)


```

Then we did the feature selection. We created a list of feature importance based on mean decreasing gini of all the features. And as we can see from the picture, we choosed the top 6 features to build this classification tree model{ap_lo (the opposite of ap_hi) is highly correlated to ap_hi, so it doesn’t gain anything to use it together.}.

```{r Classification Tree feature selection, results=TRUE}
library(randomForest)
fit_im = randomForest(cardio_tree$cardio~., data=cardio_tree)
# Create an importance based on mean decreasing gini
importance(fit_im)
varImpPlot(fit_im)
```

Next, we try to find the best depths in this model. We tried different depths and summary the results
```{r Classification Tree depths result}
loadPkg("rpart")
loadPkg("caret")



# create an empty dataframe to store the results from confusion matrices
confusionMatrixResultDf = data.frame( Depth=numeric(0), Accuracy= numeric(0), Sensitivity=numeric(0), Specificity=numeric(0), Pos.Pred.Value=numeric(0), Neg.Pred.Value=numeric(0), Precision=numeric(0), Recall=numeric(0), F1=numeric(0), Prevalence=numeric(0), Detection.Rate=numeric(0), Detection.Prevalence=numeric(0), Balanced.Accuracy=numeric(0), row.names = NULL )

for (deep in 2:8) {
  kfit <- rpart(cardio ~age+ap_lo+bmi+weight+height+cholesterol, data= cardio_tree, method="class", control = list(maxdepth = deep) )
  # 
  cm = confusionMatrix( predict(kfit, type = "class"), reference = cardio_tree[, "cardio"] ) # from caret library
  # 
  cmaccu = cm$overall['Accuracy']
  # print( paste("Total Accuracy = ", cmaccu ) )
  # 
  cmt = data.frame(Depth=deep, Accuracy = cmaccu, row.names = NULL ) # initialize a row of the metrics 
  cmt = cbind( cmt, data.frame( t(cm$byClass) ) ) # the dataframe of the transpose, with k valued added in front
  confusionMatrixResultDf = rbind(confusionMatrixResultDf, cmt)
  # print("Other metrics : ")
}

unloadPkg("caret")
```

As we can see from the results,from the depths 4, the accuracy of different depths is almost same. Therefore, we choosed the depths 6 to build the tree model.
```{r , results="asis", results=TRUE}
xkabledply(confusionMatrixResultDf, title="Cardio Classification Trees summary with varying MaxDepth")
```

Then we built this classification tree model and plot it.
```{r , echo = T, fig.dim=c(6,4), results=TRUE}
set.seed(1)
cardio_tree_fit <- rpart(cardio ~ age+ap_lo+bmi+weight+height+cholesterol , data= cardio_tree, method="class", control = list(maxdepth = 6) )

printcp(cardio_tree_fit) # display the results 
plotcp(cardio_tree_fit) # visualize cross-validation results 
summary(cardio_tree_fit) # detailed summary of splits

# plot tree 
plot(cardio_tree_fit, uniform=TRUE, main="Classification Tree for Cardio")
text(cardio_tree_fit, use.n=TRUE, all=TRUE, cex=.8)

```

```{r  }
# create attractive postcript plot of tree 
post(cardio_tree_fit, file = "cardiotreeTree2.ps", title = "Classification Tree for Cardio")
```

### Results and confusion matrix
And these are the summary results of this model and its confusion matrix.
```{r , include=T, results=TRUE}
loadPkg("caret") 
cm = confusionMatrix( predict(cardio_tree_fit, type = "class") , reference = cardio_tree[, "cardio"])
print('Overall: ')
cm$overall
print('Class: ')
cm$byClass
unloadPkg("caret")
```

```{r Classification Tree, results="asis"}
xkabledply(cm$table, "confusion matrix")
```

### Tree plot
We also tried two other ways to plot the tree, with library `rpart.plot` and a "fancy" plot using the library `rattle` to make the tree plot more beautifully. 
```{r Classification Tree fancyplot, results=TRUE}
loadPkg("rpart.plot")
rpart.plot(cardio_tree_fit)
loadPkg("rattle") 
fancyRpartPlot(cardio_tree_fit)
```

### ROC curve
Finally, we also checked the ROC and AUC of this classification tree model. As we can see, this model is kind of a good model.  
```{R ROC curve of Classification Tree, results=TRUE}
library(rpart)
rp <- rpart(cardio ~ age+ap_lo+bmi+weight+height+cholesterol, data = cardio_tree)
library(ROCR)
pred <- prediction(predict(cardio_tree_fit, type = "prob")[, 2], cardio_tree$cardio)
tree.predict.prob <- predict(cardio_tree_fit, type = "prob")[, 2]
plot(performance(pred, "tpr", "fpr"), main = "ROC Cardio")
auc = performance(pred, 'auc')
slot(auc, 'y.values')
abline(0, 1, lty = 2)

```


