---
title: "EDA"
author: "Chenxu Wang"
#date: "today"
date: "`r Sys.Date()`"
# this style requires installing rmdformats package 
output:
  html_document:
    code_folding: hide
    number_sections: true
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---
```{r init, include=FALSE}
# some of common options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
library(ezids)
library(ggplot2)
library(ggpubr)
# knitr::opts_chunk$set(warning = F, results = "markup", message = F)
knitr::opts_chunk$set(warning = F, results = "hide", message = F, echo = T)
options(scientific=T, digits = 3) 
# options(scipen=9, digits = 3) 
# ‘scipen’: integer. A penalty to be applied when deciding to print numeric values in fixed or exponential notation.  Positive values bias towards fixed and negative towards scientific notation: fixed notation will be preferred unless it is more than ‘scipen’ digits wider.
# use scipen=999 to prevent scientific notation at all times
```

# Chapter 1: Introudction of cardiovascular disease

## 1.1 Basic Information

According to the American Heart Association, cardiovascular disease can refer to a number of conditions, including heart disease, heart attack, stroke, heart failure, arrhythmia, and heart valve problems. Heart disease and stroke are the leading causes of death worldwide, ranked first and second respectively. The World Heath Organization cites unhealthy diet, physical inactivity, tobacco and alcohol use as the most important behavioral risk factors of heart disease and stroke risk. As a result, patients with cardiovascular disease often present with raised blood pressure, raised blood glucose, raised blood lipids, and BMI in the overweight or obese range.

## 1.2 The purpose of project 

Based on a report published by Centers for Disease Control and Prevention, about 610,000 people die of heart disease in the United States every year: that is 1 in every 4 deaths. 
So, in this project we aim to answer below questions by exploring data: (The smart question)

# Chapter 2: Description of Dataset
## About Dataset
The Cardiovascular Disease dataset was obtained from Kaggle   (https://www.kaggle.com/datasets/sulianova/cardiovascular-disease-dataset). There are 70000 total values (observations) in 11 features, such as age, gender, systolic blood pressure, diastolic blood pressure, and etc. The target class “cardio” equals to 1, when patient has cardiovascular disease, and it’s 0, if patient is healthy.In this data, number of people people with cardiovascular disease are 34,979 and without are 35,021. 

There are 3 types of input features:

Objective: factual information
Examination: results of medical examination
Subjective: information given by the patient.



## Variables
```{r import, results='markup'}
cardio_orig<-read.csv("cardio_train.csv", header = TRUE, sep = ";")
str(cardio_orig)
```
* `id `: ID number
* `age`: Age (in days and we will transfer to years)
* `gender`: Women(1) or Men(2)
* `height`: Height(cm)
* `weight`: Weight(kg)
* `ap_hi`: Systolic blood pressure
* `ap_lo`: Astolic blood pressure
* `cholesterol`: Cholesterol(1: normal, 2: above normal, 3: well above normal)
* `gluc`: Glucose(1: normal, 2: above normal, 3: well above normal)
* `smoke`: whether patient smokes or not(0:NO,1:Yes)
* `alco`: Alcohol intake (0:NO,1:Yes)
* `active`: Physical activity (0:NO,1:Yes)
* `cardio`:  The target class “cardio” equals to 1, when patient has cardiovascular disease, and it’s 0, if patient is healthy.

## Remove outliers
```{r removeoutliers}
cardio_remove = outlierKD2(cardio_orig, ap_hi, TRUE, TRUE, TRUE, TRUE) 
cardio_orig = outlierKD2(cardio_remove, ap_lo, TRUE, TRUE, TRUE, TRUE)
```
With minimum as low as -150 and maximum as high as 16020
Blood pressure outside the range 0-400 means person is either dead or just about to die
We have removed records with these outliers from the data


 

```{r asfactor}

cardio_orig$age=cardio_orig$age/365 # transfer days to years

cardio <- transform(cardio_orig,gender=as.factor(gender),
                      alco = as.factor(alco),
                      smoke = as.factor(smoke), 
                      active = as.factor(active), 
                      cholesterol=as.factor(cholesterol), 
                      gluc= as.factor(gluc), 
                      cardio=as.factor(cardio))
str(cardio)
```

```{r cleanNA, include=FALSE}
na_cardio<- sum(is.na(cardio))
na_cardio # there are 0 NA in cardio
cardio <- na.omit(cardio)
sum(is.na(cardio))
```

```{r results=TRUE}
str(cardio)
```
For our exploratory data analysis, we did some preprocessing including cleaning up and converting.First of all, we remove the outliers in both blood pressure columns. Then we transfer days to years for age.We dropped “NA” values from the dataset to simplify our analysis; we converted some variables into factor variables except `id`, `age`, `height`, `weight`, `ap_hi`,`ap_lo`

# Chapter 3: Categorical Variables EDA  

A brief overview of the dataset tells that 

## Present disease vs Absent disease
### What are cardiovascular disease patient look like?



### Contrast present with absent

```{r, results='markup'}

gender_1 <- ggplot(cardio, aes(x=gender, fill=cardio)) +
                   geom_bar(position="fill") +
                   scale_fill_manual(values=c("pink3", "steelblue")) +
                   scale_x_discrete(labels=c("1" = "Women", "2" = "Man")) +
                   labs(title="gender", x="", y="Percentage") +
                   theme(legend.position="top")

cholesterol_1 <- ggplot(cardio, aes(x=cholesterol, fill= cardio)) +
                   geom_bar(position="fill") +
                   scale_fill_manual(values=c("pink3", "steelblue")) +
                   scale_x_discrete(labels=c("1" = "normal", "2" = "above normal","3" = "well above normal")) +
                   labs(title="cholesterol", x="", y="Percentage") +
                   theme(legend.position="top")
gluc_1<- ggplot(cardio, aes(x=gluc, fill= cardio)) +
                   geom_bar(position="fill") +
                   scale_fill_manual(values=c("pink3", "steelblue")) +
                   scale_x_discrete(labels=c("1" = "normal", "2" = "above normal","3" = "well above normal")) +
                   labs(title="gluc", x="", y="Percentage") +
                   theme(legend.position="top")

smoke_1<- ggplot(cardio, aes(x=smoke, fill= cardio)) +
                   geom_bar(position="fill") +
                   scale_fill_manual(values=c("pink3", "steelblue")) +
                   scale_x_discrete(labels=c("0" = "No smoking", "1" = "Smoking")) +
                   labs(title="smoke", x="", y="Percentage") +
                   theme(legend.position="top")

alco_1<- ggplot(cardio, aes(x=alco, fill= cardio)) +
                   geom_bar(position="fill") +
                   scale_fill_manual(values=c("pink3", "steelblue")) +
                   scale_x_discrete(labels=c("0" = "No alcohol intake", "1" = "Alcohol intake")) +
                   labs(title="alco", x="", y="Percentage") +
                   theme(legend.position="top")

active_1<- ggplot(cardio, aes(x=active, fill= cardio)) +
                   geom_bar(position="fill") +
                   scale_fill_manual(values=c("pink3", "steelblue")) +
                   scale_x_discrete(labels=c("0" = "No physical activity", "1" = "Physical activity" )) +
                   labs(title="active", x="", y="Percentage") +
                   theme(legend.position="top")

ggarrange(gender_1, cholesterol_1,gluc_1, smoke_1, alco_1,active_1)

```
According to the plot above, we can draw the conclusion that 
*It's close to half the female and male population have cardiovascular disease.
*The higher the degree of cholesterol and glucose ， more population have this disease.
*It is surprised that whether patient smokes or not and consumes alcohol or not, Almost half of them have this disease.
* Do physical exercise can slighlt help us decrease the probability of cardiovascular disease.

# Chapter 4: Continuous Variables EDA
##4.1 KDE plot of continuous variables
The following plots are stacked for better visualization of the distribution.
```{r,results='markup'}
age_kdeplot <- ggplot(cardio, aes(x = age, fill= cardio)) + 
  geom_density(position = "stack") +
  geom_vline(aes(xintercept=mean(age)), color="black", linetype="dashed", size=1) +
  xlab("Age ") + ylab("Percentage") + 
  ggtitle(expression(atop("Age Distribution (stacked)")))+
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_fill_manual(values=c("pink3", "steelblue"),labels=c("Absent", "Present")) + 
  xlim(0, 100)
age_kdeplot

```
Above plot is stacked for better visualization of the distribution.
The percentage of disease shows that the age of patients with disease mainly concentrate on around 50.


```{r,results='markup'}
height_kdeplot<-  ggplot(cardio, aes(x = height, fill= cardio)) + 
  geom_density(position = "stack") +
  geom_vline(aes(xintercept=mean(height)), color="black", linetype="dashed", size=1) +
  xlab("Height(in cm) ") + ylab("Proportion") + 
  ggtitle(expression(atop("Height Distribution (stacked)")))+
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_fill_manual(values=c("pink3", "steelblue"),labels=c("Absent", "Present")) + 
  xlim(130, 210)

height_kdeplot
```

  
```{r,results='markup'}
weight_kdeplot<- ggplot(cardio, aes(x = weight, fill= cardio)) + 
  geom_density(position = "stack") +
  geom_vline(aes(xintercept=mean(weight)), color="black", linetype="dashed", size=1) +
  xlab("Weight(in kg) ") + ylab("Proportion") + 
  ggtitle(expression(atop("Weight Distribution (stacked)")))+
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_fill_manual(values=c("pink3", "steelblue"),labels=c("Absent", "Present")) + 
  xlim(0, 200)

weight_kdeplot
```



```{r,results='markup'}
ap_hi_kdeplot<- ggplot(cardio, aes(x = ap_hi, fill= cardio)) + 
  geom_density(position = "stack") +
  geom_vline(aes(xintercept=mean(ap_hi)), color="black", linetype="dashed", size=1) +
  xlab("Systolic blood pressure(ap_hi) ") + ylab("Proportion") + 
  ggtitle(expression(atop("Systolic blood pressure Distribution (stacked)")))+
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_fill_manual(values=c("pink3", "steelblue"),labels=c("Absent", "Present")) + 
  xlim(50, 250)

ap_hi_kdeplot
```


```{r,results='markup'}
ap_lo_kdeplot<-ggplot(cardio, aes(x = ap_lo, fill= cardio)) + 
  geom_density(position = "stack") +
  geom_vline(aes(xintercept=mean(ap_lo)), color="black", linetype="dashed", size=1) +
  xlab("Diastolic blood pressure(ap_lo)) ") + ylab("Proportion") + 
  ggtitle(expression(atop("Diastolic blood pressure Distribution (stacked)")))+
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_fill_manual(values=c("pink3", "steelblue"),labels=c("Absent", "Present")) + 
  xlim(50, 150)

ap_lo_kdeplot
```




