---
title: "Heartbreakers: Estimating the Risk Factors of Cardiovascular Disease"
author: "By: Hovhannes Arestakesyan, Karina Martinez, Timberley Thompson, Chenxu Wang, & Zhe Yu"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r init, include=FALSE}
#RMD Settings
knitr::opts_chunk$set(warning = F, results = "hide", message = F)
options(scientific=T, digits = 3) 

#Packages
library(ezids)
library(tidyverse)
library(tinytex)
require(rmarkdown)

#Data
cardio_orig = data.frame(read.csv("cardio_train.csv", header = TRUE, sep=';'))
str(cardio_orig)

cardio = cardio_orig[-c(1)] #get rid of id
cardio$age = cardio$age/365 # transfer days to years
cardio$bmi = cardio$weight/((cardio$height/100)**2) #add bmi variable

cardio_clean = outlierKD2(cardio, bmi, TRUE, TRUE, TRUE, TRUE) 
cardio_clean = outlierKD2(cardio_clean, ap_hi, TRUE, TRUE, TRUE, TRUE) 
cardio_clean = outlierKD2(cardio_clean, ap_lo, TRUE, TRUE, TRUE, TRUE)
cardio_clean = na.omit(cardio_clean) #remove NA rows

cardio_clean_final = cardio_clean
cardio_clean_final$gender = as.factor(cardio_clean$gender)
cardio_clean_final$cholesterol = as.factor(cardio_clean$cholesterol) 
cardio_clean_final$gluc = as.factor(cardio_clean$gluc)
cardio_clean_final$smoke = as.factor(cardio_clean$smoke)
cardio_clean_final$alco = as.factor(cardio_clean$alco)
cardio_clean_final$active = as.factor(cardio_clean$active)
cardio_clean_final$cardio = as.factor(cardio_clean$cardio)
```
# I. Background

Cardiovascular disease is "the leading cause of death globally," killing an estimated 17.9 million people each year (World Health Organization). Cardiovascular disease (also known as heart disease) refers to any of a group of conditions affecting one's heart and blood vessels; these conditions can include heart failure, arrhythmia, aortic disease and more (Cleveland Clinic).

The World Health Organization asserts that an "unhealthy diet, physical inactivity, tobacco use" and alcohol abuse are the most important behavioral risk factors in developing heart disease. Engaging in such activities can often lead to "raised blood pressure, raised blood glucose... and obesity" (World Health Organization).

This research paper is an investigation of The World Health Organization's statements. The research team analyzes data on cardiovascular disease to determine the risk factors of the disease and offer recommendations to reduce risk.

## SMART Questions
Specifically, the research team answers the following Specific, Measurable, Achievable, Relevant, and Time-oriented (SMART) questions:  

  1. How accurately can the model predict the probability that a patient has cardiovascular disease?
  2. What are the top three factors associated with cardiovascular disease? Does this corroborate the risk factors observed by the World Health Organization?  
  3. Is there any association between an individual's sex and whether they have 
cardiovascular disease ? Are men more prone to cardiovascular diseases as compared to
women?
  4. What recommendations can be made to reduce cardiovascular risk and prevent heart attacks and strokes?  


## Dataset Description

Data analysis was performed on the Cardiovascular Disease dataset from [Kaggle.com](https://www.kaggle.com/datasets/sulianova/cardiovascular-disease-dataset). After some data cleaning, the final dataset, `cardio_clean_final`, has `r nrow(cardio_clean_final)` observations and the following variables:

### Variables

Variable | Type | Definition  
  :-     |  :-- |  :----
`id`	 | Character | ID number
`age`	 | Number | Age in years
`gender` |Factor w/ 2 levels | Gender (1: women, 2: men)
`height` | Number | Height in centimeters (cm)
`weight` |Number | Weight in kilograms (kg)
`bmi` | Number | Body Max Index (kg/m^2^)
`ap_hi` | Integer | Systolic blood pressure
`ap_lo` | Integer | Diastolic blood pressure
`cholesterol` |Factor w/ 3 levels| Cholesterol (1: normal, 2: above normal, 3: well above normal)
`gluc` |Factor w/ 3 levels| Glucose (1: normal, 2: above normal, 3: well above normal)
`smoke` |Factor w/ 2 levels| Whether patient smokes or not (0: no, 1: yes)
`alco` |Factor w/ 2 levels| Whether patient consumes alcohol or not (0: no, 1: yes)
`active` |Factor w/ 2 levels| Whether patient is physically active or not (0: no, 1: yes)
`cardio` |Factor w/ 2 levels| Presence or absence of cardiovascular disease (0: absent, 1: present)

# II. Exploratory Data Analysis
## Categorical Variables

```{r cat.graphs, results='markup'}
require(patchwork)

#cholesterol
cholesterol.1 <- ggplot(cardio_clean_final, aes(x=cardio, fill= cholesterol)) +
  geom_bar(position="fill") +
  scale_x_discrete(labels=c("1" = "Heart Disease", "0" = "Healthy")) +
  scale_y_continuous(labels = scales::percent) +
  labs( title = "Cholesterol", y = "", x = "",
        fill = "Cholesterol Level") +
  scale_fill_manual(values=c("#f4cccc", "#e06666", "#990000")) +
  theme_minimal()

#Glucose
gluc.1 <- ggplot(cardio_clean_final, aes(x=cardio, fill= gluc)) +
  geom_bar(position="fill") +
  scale_x_discrete(labels=c("1" = "Heart Disease", "0" = "Healthy")) +
  scale_y_continuous(labels = scales::percent) +
  labs( title = "Glucose", y = "", x = "",
        fill = "Glucose Level") +
  scale_fill_manual(values=c("#f4cccc", "#e06666", "#990000")) +
  theme_minimal()

#Gender
cardio_clean_final$gender.yn<- as.factor(ifelse(cardio_clean_final$gender== "1", "Women", "Men"))

gender.1 <- ggplot(cardio_clean_final, aes(x=cardio, fill= gender.yn)) +
  geom_bar(position="fill") +
  scale_x_discrete(labels=c("1" = "Heart Disease", "0" = "Healthy")) +
  scale_y_continuous(labels = scales::percent) +
  labs( title = "Gender",
        y = "", x = "", fill = "") +
  scale_fill_manual(values=c("#0000da", "#990000")) +
  theme_minimal()

#Smoke
cardio_clean_final$smoke.yn<- as.factor(ifelse(cardio_clean_final$smoke== "1", "Yes", "No"))

smoke.1 <- ggplot(cardio_clean_final, aes(x=cardio, fill= smoke.yn)) +
  geom_bar(position="fill") +
  scale_x_discrete(labels=c("1" = "Heart Disease", "0" = "Healthy")) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Tobacco Use", y = "", x = "", fill = "Smoker") +
  scale_fill_manual(values=c("#0000da", "#990000")) +
  theme_minimal()

#alcohol
  cardio_clean_final$alco.yn<- as.factor(ifelse(cardio_clean_final$alco== "1", "Yes", "No"))

alco.1 <- ggplot(cardio_clean_final, aes(x=cardio, fill= alco.yn)) +
  geom_bar(position="fill") +
  scale_x_discrete(labels=c("1" = "Heart Disease", "0" = "Healthy")) +
  scale_y_continuous(labels = scales::percent) +
  labs( title = "Alcohol Use", y = "", x = "", fill = "Drinks Alcohol") +
  scale_fill_manual(values=c("#0000da", "#990000")) +
  theme_minimal()

#active
cardio_clean_final$active.yn<- as.factor(ifelse(cardio_clean_final$active== "1", "Yes", "No"))

active.1 <- ggplot(cardio_clean_final, aes(x=cardio, fill= active.yn)) +
  geom_bar(position="fill") +
  scale_x_discrete(labels=c("1" = "Heart Disease", "0" = "Healthy")) +
  scale_y_continuous(labels = scales::percent) +
  labs( title = "Physical Activity", y = "", x = "", fill = "Physically Active") +
  scale_fill_manual(values=c("#0000da", "#990000")) +
  theme_minimal() 

gender.1 + cholesterol.1 + gluc.1 + smoke.1 +  alco.1 + active.1 + plot_layout(ncol = 2)
```

According to the graphs, how do CVD patients differ from non-CVD  patients with respect to the categorical variables? 

## Chi-Squared Test
What was the purpose of the chi-squared test?

Categorical Variable | Chi-Square | P-Value | df  
  :-     |  :- |  :-| :-
  `gender`  | 0.4 | 0.5 | 1
  `cholesterol` | 2991 | <2e-16 | 2
  `glucose` | 478 | <2e-16 | 2
  `smoke` | 21 | 4e-06 | 1
  `alco` | 7 | 0.006 | 1
  `active` | 81 | <2e-16 | 1
```{r chi}
#Gender
cardio_vs_gender = table(cardio_clean_final$cardio, cardio_clean_final$gender)
chi_gender = chisq.test(cardio_vs_gender)

#Cholesterol
cardio_vs_cholesterol=table(cardio_clean_final$cardio, cardio_clean_final$cholesterol)
chi_cholesterol = chisq.test(cardio_vs_cholesterol)

#Glucose
cardio_vs_glucose = table(cardio_clean_final$cardio, cardio_clean_final$gluc)
chi_glucose = chisq.test(cardio_vs_glucose)

#Smoking
cardio_vs_smoke = table(cardio_clean_final$cardio, cardio_clean_final$smoke)
chi_smoke = chisq.test(cardio_vs_smoke)

#Alcohol
cardio_vs_alco = table(cardio_clean_final$cardio, cardio_clean_final$alco)
chi_alco = chisq.test(cardio_vs_alco)

#Physical Activity
cardio_vs_active = table(cardio_clean_final$cardio, cardio_clean_final$active)
chi_active = chisq.test(cardio_vs_active)


```

In plain English, describe what the results of the chi-squared test reveal. What conclusions can be drawn from the results? Do they corroborate the conclusions drawn from the graphs?

## Continuous Variables

```{r cont.graphs}
disease <- subset(cardio_clean_final, cardio==1)
healthy <- subset(cardio_clean_final, cardio==0)

#age
cardio_clean_final$cardio.yn<- as.factor(ifelse(cardio_clean_final$active== "1", "Yes", "No"))

age.kde <- ggplot(cardio_clean_final, aes(x = age, fill= cardio.yn)) + 
  geom_density(position = "stack") +
  geom_vline(aes(xintercept=mean(disease$age)), color="pink3", linetype="dashed", size=1) +
  geom_vline(aes(xintercept=mean(healthy$age)), color="steelblue", linetype="dashed", size=1) +
  scale_x_continuous(limits = c(30,70)) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Age",
       subtitle = "Age Distribution (Stacked)",
       fill = "Has CVD", x = "Age", y = " ") +
  scale_fill_manual(values=c("#0000da", "#990000")) +
  theme_minimal()


#height
height.kde <- ggplot(cardio_clean_final, aes(x = height, fill= cardio.yn)) + 
  geom_density(position = "stack") +
  geom_vline(aes(xintercept=mean(disease$height)), color="pink3", linetype="dashed", size=1) +
  geom_vline(aes(xintercept=mean(healthy$height)), color="steelblue", linetype="dashed", size=1) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Height",
       subtitle = "Height Distribution (Stacked)",
       fill = "Has CVD", x = "Height (cm)", y = " ") +
  scale_fill_manual(values=c("#0000da", "#990000")) +
  theme_minimal()



#weight
weight.kde <- ggplot(cardio_clean_final, aes(x = weight, fill= cardio.yn)) + 
  geom_density(position = "stack") +
  geom_vline(aes(xintercept=mean(disease$weight)), color="pink3", linetype="dashed", size=1) +
  geom_vline(aes(xintercept=mean(healthy$weight)), color="steelblue", linetype="dashed", size=1) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Weight",
       subtitle = "Weight Distribution (Stacked)",
       fill = "Has CVD", x = "Weight (kg)", y = " ") +
  scale_fill_manual(values=c("#0000da", "#990000")) +
  theme_minimal()



#bmi
bmi.kde <- ggplot(cardio_clean_final, aes(x = bmi, fill= cardio.yn)) + 
  geom_density(position = "stack") +
  geom_vline(aes(xintercept=mean(disease$bmi)), color="pink3", linetype="dashed", size=1) +
  geom_vline(aes(xintercept=mean(healthy$bmi)), color="steelblue", linetype="dashed", size=1) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "BMI",
       subtitle = "BMI Distribution (Stacked)",
       fill = "Has CVD", x = "BMI", y = " ") +
  scale_fill_manual(values=c("#0000da", "#990000")) +
  theme_minimal()



#systolic bp
ap_hi.kde <- ggplot(cardio_clean_final, aes(x = ap_hi, fill= cardio.yn)) + 
  geom_density(position = "stack") +
  geom_vline(aes(xintercept=mean(disease$ap_hi)), color="pink3", linetype="dashed", size=1) +
  geom_vline(aes(xintercept=mean(healthy$ap_hi)), color="steelblue", linetype="dashed", size=1) +
  scale_y_continuous(labels = scales::percent) +
  scale_x_continuous(breaks = c(seq(90,170,20))) +
  labs(title = "Systolic BP",
       subtitle = "Systolic BP Distribution (Stacked)",
       fill = "Has CVD", x = "Systolic BP", y = " ") +
  scale_fill_manual(values=c("#0000da", "#990000")) +
  theme_minimal()



#diastolic bp
ap_lo.kde <- ggplot(cardio_clean_final, aes(x = ap_lo, fill= cardio.yn)) + 
  geom_density(position = "stack") +
  geom_vline(aes(xintercept=mean(disease$ap_lo)), color="pink3", linetype="dashed", size=1) +
  geom_vline(aes(xintercept=mean(healthy$ap_lo)), color="steelblue", linetype="dashed", size=1) +
  scale_y_continuous(labels = scales::percent) +
  scale_x_continuous(breaks = c(seq(65,105,10))) +
  labs(title = "Diastolic BP",
       subtitle = "Diastolic BP Distribution (Stacked)",
       fill = "Has CVD", x = "Diastolic BP", y = " ") +
  scale_fill_manual(values=c("#0000da", "#990000")) +
  theme_minimal()



require(ggpubr)
ggarrange(age.kde, height.kde, weight.kde, bmi.kde, ap_hi.kde, ap_lo.kde, ncol=2, nrow=3, common.legend = TRUE, legend = "bottom", heights=c(2,2,2))


```

According to the graphs, how do CVD patients differ from non-CVD  patients with respect to the continuous variables? Please note that the pink dashed lines indicate the means for CVD patients while the blue dashed lines indicate the means for healthy patients.

## Welch Two-Sample T-Test
What is the purpose of the  t-test? 

Continuous Variable | CVD Avg. | Healthy Avg. | t | p-value
:- | :- | :- | :- | :-
`age` | 55 | 51.8 | 61 | <2e-16
`height` | 165 | 165 | -4 | 1e-04
`weight`| 75.3 | 71.3 | 41 | <2e-16
`bmi` | 27.8 | 26.3 | 46 | <2e-16
`ap_hi` | 133 | 120 | 120 | <2e-16
`ap_lo` | 84.3 | 79.1 | 89 | <2e-16

```{r ttest}

cardio_pos = subset(cardio_clean_final, cardio_clean_final$cardio == 1)
cardio_neg = subset(cardio_clean_final, cardio_clean_final$cardio == 0)

#age
ttest_age = t.test(cardio_pos$age, cardio_neg$age)
ttest_age

#height
ttest_height = t.test(cardio_pos$height, cardio_neg$height)
ttest_height

#weight
ttest_weight = t.test(cardio_pos$weight, cardio_neg$weight)
ttest_weight

#bmi
ttest_bmi = t.test(cardio_pos$bmi, cardio_neg$bmi)
ttest_bmi

#ap.hi
ttest_ap_hi = t.test(cardio_pos$ap_hi, cardio_neg$ap_hi)
ttest_ap_hi

#ap.lo
ttest_ap_lo = t.test(cardio_pos$ap_lo, cardio_neg$ap_lo)
ttest_ap_lo

```
In plain English, describe what the results of the t-test reveal. What conclusions can be drawn from the results? 


## Correlations
What is the purpose of examining correlations?

```{r corrplot}
loadPkg("corrplot")


# corrplot with spearman method for categorical variables
cardio.cor.1 <- cor(cardio_clean, method="spearman")
corrplot(cardio.cor.1, order = "hclust", type="lower", addCoef.col="darkred", number.cex=0.6, tl.cex=1,title="Correlation Matrix", mar=c(0,0,1,0))
```

What does the corrplot tell us?

** further exploration of correlations * *

# III. Model Comparison

## Feature selection  
In the EDA, t-tests and Chi-square tests were used to compare whether the response variables `age`, `gender`, `height`, `weight`, `ap_hi`, `ap_lo`, `cholesterol`, `gluc`, `smoke`, `alco`, `active` and `bmi` have effects on the predictor variable `cardio`, and the p-values of the tests were shown in the table below:  
```{r results='markup'}
tab <- matrix(c( '<2e-16','0.6', '8e-07', '<2e-16', '<2e-16', '<2e-16', '<2e-16', '<2e-16', '1e-06', '0.006', '<2e-16', '<2e-16'), ncol=12, byrow = TRUE)
rownames(tab) <- "p_value"
colnames(tab) <- c("Age_year", "Gender", "Height","Weight","Systolic(ap_hi)", "Diastolic(ap_low)", "Cholesterol","Glucose","smoke", "Alcohol", "Physical_activity","BMI")
tab <- as.table(tab)
xkabledply(tab, "p-value Comparison")  
```

**From the above table, we found that all the response variables have relationship with the predictor variable `cardio` except the `gender` variable. Next we further confirmed this using the `bestglm` function.  

```{r bestglm, results='markup'}
loadPkg("bestglm")
loadPkg("leaps")
loadPkg("ggplot2")
cardio_clean_final_glm <- cardio_clean_final
colnames(cardio_clean_final_glm)[12] <- "y"
cardio_clean_final_glm <- cardio_clean_final_glm[c("age", "gender", "height", "weight","ap_hi", "ap_lo", "cholesterol", "gluc", "smoke", "alco", "active", "bmi", "y")]

res.bestglm <- bestglm(Xy = cardio_clean_final_glm, family = binomial,
            IC = "AIC", method = "exhaustive")
summary(res.bestglm)
res.bestglm$BestModels
summary(res.bestglm$BestModels)
unloadPkg("bestglm") 
unloadPkg("leaps") 
```  
`gender` and `bmi` were excluded from the model that has lowest criterion value, indicating these two valuables may not have strong effects on `cardio`. But among the 5 models the function generated, the criterion values are very close, and `gender` was voted as `FALSE` three time, while `bmi` was only voted twice. So we are not confidence in removing these two valuables from the following modeling steps. In the KNN modeling, we compare whether these two variables can affect the model performances. Also, since the correlations matrix showed the valuables `age`, `cholesterol`, `bmi`, `ap_hi` and `ap_lo` had relatively high coefference values with `cardio`, we also performed the KNN modeling with these 5 valuables.    

## K-Nearest Neighbor (KNN)     
### Convert factors into numeric, and Scale/center the data       
```{r results='markup'}
cardio_num <- cardio_clean_final_glm
colnames(cardio_num)[13] <- "cardioi"
cardio_num[1:12] <- sapply(cardio_num[1:12],as.numeric)
cardio_scale <- cardio_num
cardio_scale[1:12] <- as.data.frame(scale(cardio_num[1:12], center = TRUE, scale = TRUE))
str(cardio_scale)
```  
Dataframe structure after pre-modeling treatment was shown in the above table. All the response variables were converted in to numeric, and were scaled and centered. They are ready for KNN modeling.  

### KNN modeling using variables with all 13 variables (Model 1)  
#### Train / Test split:  
The dataframe was split into training subset (70% of the total dataframe) and test subset (30% of the total dataframe).  
```{r}
set.seed(42)
cardio_sample <- sample(2, nrow(cardio_scale), replace=TRUE, prob=c(0.7, 0.3))
cardio_training_x <- cardio_scale[cardio_sample==1, 1:12]
cardio_test_x <- cardio_scale[cardio_sample==2, 1:12]
cardio_training_y <- cardio_scale[cardio_sample==1, 13]
cardio_test_y <- cardio_scale[cardio_sample==2, 13]
```

#### Looking for the best K value:  
Odd numbers from 3 to 40 were test one by one to get the best K value. The accuracy values from the model that used these K values were shown in the table and the line graph. According the results, K = 19 is the best K value for Model 1.  

```{r}
loadPkg("FNN")
loadPkg("gmodels")
loadPkg("caret")
K_test <- seq(3, 40, by = 2)
ResultDf = data.frame()
for (k_v in K_test) {
  cardio_pred <- knn(train = cardio_training_x, test = cardio_test_x, cl=cardio_training_y, k=k_v)
  
  cm = confusionMatrix(cardio_pred, reference = cardio_test_y )
  
  cmaccu = cm$overall['Accuracy']

  cmt = data.frame(k=k_v, Total.Accuracy = cmaccu, row.names = NULL ) 
  
  ResultDf = rbind(ResultDf, cmt)
}
```
```{r results='markup'}
xkabledply(ResultDf, "Total Accuracy Summary for KNN Model 1")
```

```{r}
library(ggplot2)
ggplot(ResultDf, aes(x=k, y=Total.Accuracy)) +
  geom_line(color = "skyblue", size = 1.5) +
  geom_point(size = 3, color = "blue") + 
  labs(title = "Total.Accuracy vs K (Model 1)")
```  
#### Modeling results with the best K value (Model 1, K = 19)  

Detailed results and the confusion matrix using K=19 were shown below.  

```{r results='markup'}
   cardio_pred_19 <- knn(train = cardio_training_x, test = cardio_test_x, cl=cardio_training_y, k=19)
  KNN_Mode_1 = confusionMatrix(cardio_pred_19, reference = cardio_test_y)
  xkabledply(as.matrix(KNN_Mode_1), title = paste("Confusion Matrix for KNN Model 1") ) 
  xkabledply(data.frame(KNN_Mode_1$overall['Accuracy']), title=paste("Total Accuracy of KNN Model 1"))
  xkabledply(data.frame(KNN_Mode_1$byClass), title=paste("Metrics of KNN Model 1"))
```  
 The overall performance is acceptable, but we still want to see whether remove `gender` and `bmi` can improve the modeling.  

### KNN modeling using variables without `gender` and `bmi` (Model 2)  
#### Train / Test split:  
The dataframe was split into training subset (70% of the total dataframe) and test subset (30% of the total dataframe). The valuables `gender` and `bmi` were excluded from the train/test split.
```{r}
cardio_scale2 <- cardio_scale[c(1, 3:11, 13)]
cardio_scale2
set.seed(42)
cardio_sample2 <- sample(2, nrow(cardio_scale2), replace=TRUE, prob=c(0.7, 0.3))
cardio_training_x2 <- cardio_scale2[cardio_sample2==1, 1:10]
cardio_test_x2 <- cardio_scale2[cardio_sample2==2, 1:10]
cardio_training_y2 <- cardio_scale2[cardio_sample2==1, 11]
cardio_test_y2 <- cardio_scale2[cardio_sample2==2, 11]
```

#### Looking for the best K value:  
Odd numbers from 3 to 40 were used for the test. The accuracy values from the model that used these K values were shown in the table and the line graph. According the results, K = 29 is the best K value for the KNN modeling without `gender` and `bmi`.

```{r}
K_test <- seq(3, 40, by = 2)
ResultDf2 = data.frame()
for (k_v in K_test) {
  cardio_pred2 <- knn(train = cardio_training_x2, test = cardio_test_x2, cl=cardio_training_y2, k=k_v)
  
  cm2 = confusionMatrix(cardio_pred2, reference = cardio_test_y2 )
  
  cmaccu2 = cm2$overall['Accuracy']

  cmt2 = data.frame(k=k_v, Total.Accuracy = cmaccu2, row.names = NULL ) 
  
  ResultDf2 = rbind(ResultDf2, cmt2)
}
```
```{r results='markup'}
xkabledply(ResultDf2, "Total Accuracy Summary for KNN Model 2")
```

```{r}
library(ggplot2)
ggplot(ResultDf2, aes(x=k, y=Total.Accuracy)) +
  geom_line(color = "skyblue", size = 1.5) +
  geom_point(size = 3, color = "blue") + 
  labs(title = "Total.Accuracy vs K (Model 2)")
```  

#### Modeling results with the best K value (Model 2, K = 29)  

Detailed results and the confusion matrix of model 2 using k = 29 were shown below. Removing `gender` and `bmi` slightly improved the performance. 

```{r results='markup'}
  cardio_pred_29 <- knn(train = cardio_training_x2, test = cardio_test_x2, cl=cardio_training_y2, k=29)
  KNN_Mode_2 = confusionMatrix(cardio_pred_29, reference = cardio_test_y2)
  xkabledply(as.matrix(KNN_Mode_2), title = paste("Confusion Matrix for KNN Model 2") ) 
  xkabledply(data.frame(KNN_Mode_2$overall['Accuracy']), title=paste("Total Accuracy of KNN Model 2"))
  xkabledply(data.frame(KNN_Mode_2$byClass), title=paste("Metrics of KNN Model 2"))
```  

Next we want to check whether only use the 5 valuables that provided by correlation matrix will be a better model.  

### KNN modeling using only `age`, `cholesterol`, `bmi`, `ap_hi` and `ap_lo` (Model 3)  
#### Train / Test split:  
The dataframe was split into training subset (70% of the total dataframe) and test subset (30% of the total dataframe). Only `age`, `cholesterol`, `bmi`, `ap_hi` and `ap_lo` were used for the modeling. 
```{r}
cardio_scale3 <- cardio_scale[c(1, 5:7, 12:13)]
cardio_scale3
set.seed(42)
cardio_sample3 <- sample(2, nrow(cardio_scale3), replace=TRUE, prob=c(0.7, 0.3))
cardio_training_x3 <- cardio_scale3[cardio_sample3==1, 1:5]
cardio_test_x3 <- cardio_scale3[cardio_sample3==2, 1:5]
cardio_training_y3 <- cardio_scale3[cardio_sample3==1, 6]
cardio_test_y3 <- cardio_scale3[cardio_sample3==2, 6]
```

#### Looking for the best K value:  
Odd numbers from 3 to 40 were used for the test. The accuracy values from the model that used these K values were shown in the table and the line graph. According the results, K = 31 is the best K value for the KNN modeling.

```{r}
K_test <- seq(3, 40, by = 2)
ResultDf3 = data.frame()
for (k_v in K_test) {
  cardio_pred3 <- knn(train = cardio_training_x3, test = cardio_test_x3, cl=cardio_training_y3, k=k_v)
  
  cm3 = confusionMatrix(cardio_pred3, reference = cardio_test_y3 )
  
  cmaccu3 = cm3$overall['Accuracy']

  cmt3 = data.frame(k=k_v, Total.Accuracy = cmaccu3, row.names = NULL ) 
  
  ResultDf3 = rbind(ResultDf3, cmt3)
}
```
```{r results='markup'}
xkabledply(ResultDf3, "Total Accuracy Summary")
```

```{r}
library(ggplot2)
ggplot(ResultDf3, aes(x=k, y=Total.Accuracy)) +
  geom_line(color = "skyblue", size = 1.5) +
  geom_point(size = 3, color = "blue") + 
  labs(title = "Total.Accuracy vs K")
```  

#### Modeling results with the best K value (Model 3, K = 31)  

The confusion matrix of model 3 using k = 31 were shown below. 

```{r results='markup'}
  cardio_pred_31 <- knn(train = cardio_training_x3, test = cardio_test_x3, cl=cardio_training_y3, k=31)
  KNN_Mode_3 = confusionMatrix(cardio_pred_31, reference = cardio_test_y3 )
  xkabledply(as.matrix(KNN_Mode_3), title = paste("Confusion Matrix for KNN Model 3") )
```  

```{r results='markup'}
 xkabledply(data.frame(KNN_Mode_1$overall['Accuracy'], KNN_Mode_2$overall['Accuracy'], KNN_Mode_3$overall['Accuracy']), title=paste("Total Accuracy Comparision of the Three KNN Models"))

 xkabledply(data.frame(KNN_Mode_1$byClass, KNN_Mode_2$byClass, KNN_Mode_3$byClass), title=paste("Metrics Comparision of KNN the Three KNN Models"))
```
**The total accuracy and other metrics of the three models were listed in the above tables. For summary, all three models have an acceptable performance, but not outstanding. More specifically, model 2 has a overall better performance comparing to model 1. But model 2 and model 3 have their own strength and weaknesses. The accuracy and recall value of model 2 are higher than model 3, while the precision value is higher in model 3. Recall value is more related to false negative rate and the precision value is used to represent to the false positive rate. It's hard to say which model is the best one, but we should choose the proper model according to our aims. If minimizing false positives is our focus, model 3 might be a good option, and model 2 is more suitable is we want to minimize  false negative rate.**  
**One of the limitation of the KNN modeling is this model has difficulties in identifying feature importance. So from here we can't answer the question that which are the top 3 factors that affect the onset of cardiovascular diseases. Our following modeling using logistic regression and classification tree will answer the question.**  


## Classification Tree
How did the EDA influence your choices when developing this model? (For example: Why is a classification tree appropriate for this data? What variables did you use and why?) What is the purpose of the model?

--insert graphic with results--

What do the results tell us?

### Model Evaluation
...

**Conclusion:** Is this a good model? Why or why not? What are its strengths and weaknesses? How does it compare to the logistic and KNN models?


# IV. Conclusion
**1. How accurately can the model predict the probability that a patient has cardiovascular disease?**  
Provide an answer. Use the results of our analysis to support the answer.
  
**2. What are the top three factors associated with cardiovascular disease? Does this corroborate the risk factors observed by the World Health Organization?**  
Provide an answer. Use the results of our analysis to support the answer.

**3. Is there an association between and individual's sex and whether they have cardiovascular disease?**  
Provide an answer. Use the results of our analysis to support the answer.

**4. What recommendations can be made to reduce cardiovascular risk and prevent heart attacks and strokes?**  
Provide an answer. Use the results of our analysis to support the answer.

Discuss the limitations of the dataset and our analysis. Discuss what can be done to improve research in this area.

# References
World Health Organization. (n.d.). *Cardiovascular diseases*. World Health Organization. Retrieved August 2, 2022, from https://www.who.int/health-topics/cardiovascular-diseases#tab=tab_1 

Cleveland Clinic medical professional (Ed.). (2021, October 5). *Cardiovascular disease: Heart disease causes and symptoms.* Cleveland Clinic. Retrieved August 2, 2022, from https://my.clevelandclinic.org/health/diseases/21493-cardiovascular-disease#:~:text=Cardiovascular%20disease%20is%20the%20leading,women%20die%20from%20cardiovascular%20disease. 









