---
title: "Team 2 Technical Analysis"
author: "Hovhannes Arestakesyan, Karina Martinez, Timberley Thompson, Chenxu Wang, & Zhe Yu"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r setup, include=FALSE}
#Libraries
library(ezids)
library(tidyverse)

#Knit Options
knitr::opts_chunk$set(warning = F, results = "hide", message = F)
options(scientific=T, digits = 3) 
```

# Cardiovascular Dataset
```{r}
#Import Data
cardio_orig = data.frame(read.csv("cardio_train.csv", header = TRUE, sep=';'))
str(cardio_orig)
```
This Cardiovascular Disease dataset is from [Kaggle](https://www.kaggle.com/datasets/sulianova/cardiovascular-disease-dataset). The variables are:  

Variable |  Definition  
  :-     |  :---- 
id	 | ID number
age	 | Age in days
gender | Gender (1: women, 2: men)
height | Height in centimeters (cm)
weight | Weight in kilograms (kg)
ap_hi | Systolic blood pressure
ap_lo | Diastolic blood pressure
cholesterol | Cholesterol (1: normal, 2: above normal, 3: well above normal)
gluc | Glucose (1: normal, 2: above normal, 3: well above normal)
smoke | Whether patient smokes or not (0: no, 1: yes)
alco | Whether patient consumes alcohol or not (0: no, 1: yes)
active | Whether patient is physically active or not (0: no, 1: yes)
cardio | Presence or absence of cardiovascular disease (0: absent, 1: present)

There are 70000 observations in the dataset.

# Data Cleanup
```{r}
#Fix Variables
cardio = cardio_orig[-c(1)] #get rid of id
cardio$age = cardio$age/365 # transfer days to years
cardio$gender = as.factor(cardio$gender)
cardio$cholesterol = as.factor(cardio$cholesterol) 
cardio$gluc = as.factor(cardio$gluc)
cardio$smoke = as.factor(cardio$smoke)
cardio$alco = as.factor(cardio$alco)
cardio$active = as.factor(cardio$active)
cardio$cardio = as.factor(cardio$cardio)
cardio$bmi = cardio$weight/((cardio$height/100)**2) #add bmi variable

summary(cardio)
```

We remove the `id` column and converted `age` from days to years.  We added a new column `bmi`, for body mass index. Body mass index (BMI) is a measure of body fat based on height and weight, and the formula is $BMI = kg/m^2$. We also converted categorical variables from integer to factor. The new data frame is saved as `cardio`. 

We see that there are unexpected minimum and maximum observations for `height`, `weight`, `ap_hi`, `ap_lo` and `bmi`.  Blood pressure measurements as low as -150 and as high as 16020 are incompatable with life, as are BMIs of 3.5 and 299.  We use the `outlierKD2()` and `na.omit()` functions to remove the observations with outliers. 

```{r}
#Remove Outliers
cardio_clean = outlierKD2(cardio, bmi, TRUE, TRUE, TRUE, TRUE) 
cardio_clean = outlierKD2(cardio_clean, ap_hi, TRUE, TRUE, TRUE, TRUE) 
cardio_clean = outlierKD2(cardio_clean, ap_lo, TRUE, TRUE, TRUE, TRUE)

cardio_clean_final = na.omit(cardio_clean) #remove NA rows
```

After the outliers were dropped, the QQ-plots improved, indicating the distribution is closer to normal. The "stair-case" pattern on the blood pressure QQ-plots is due to discrete or rounded values. We can also see the more normal distribution in the histograms and box plots for all of the variables.  
The new dataframe is saved as `cardio_clean_final`. There are `r nrow(cardio_clean_final)` observations in the new dataset.

```{r}
xkablesummary(cardio_clean_final, title = "Clean Cardio Data Frame Summary" )
```

# Exploratory Data Analysis (EDA)

```{r}
disease <- subset(cardio_clean_final, cardio==1)
healthy <- subset(cardio_clean_final, cardio==0)
summary(disease)
summary(healthy)

```

## Categorical Variables
A brief overview of the dataset tells us we have the following independent categorical variables: `gender`, `cholesterol`, `gluc`, `smoke`, and `active`.

### Graphics
```{r, results='markup'}
require(patchwork)

#cholesterol
cholesterol.1 <- ggplot(cardio_clean_final, aes(x=cardio, fill= cholesterol)) +
  geom_bar(position="fill") +
  scale_x_discrete(labels=c("1" = "Heart Disease", "0" = "Healthy")) +
  scale_y_continuous(labels = scales::percent) +
  labs( title = "Cholesterol", y = "", x = "",
        fill = "Cholesterol Level") +
  scale_fill_manual(values=c("#f4cccc", "#e06666", "#990000")) +
  theme_minimal()

#Glucose
gluc.1 <- ggplot(cardio_clean_final, aes(x=cardio, fill= gluc)) +
  geom_bar(position="fill") +
  scale_x_discrete(labels=c("1" = "Heart Disease", "0" = "Healthy")) +
  scale_y_continuous(labels = scales::percent) +
  labs( title = "Glucose", y = "", x = "",
        fill = "Glucose Level") +
  scale_fill_manual(values=c("#f4cccc", "#e06666", "#990000")) +
  theme_minimal()

#Gender
cardio_clean_final$gender.yn<- as.factor(ifelse(cardio_clean_final$gender== "1", "Women", "Men"))

gender.1 <- ggplot(cardio_clean_final, aes(x=cardio, fill= gender.yn)) +
  geom_bar(position="fill") +
  scale_x_discrete(labels=c("1" = "Heart Disease", "0" = "Healthy")) +
  scale_y_continuous(labels = scales::percent) +
  labs( title = "Gender",
        y = "", x = "", fill = "") +
  scale_fill_manual(values=c("#990000", "#0000da")) +
  theme_minimal()

#Smoke
cardio_clean_final$smoke.yn<- as.factor(ifelse(cardio_clean_final$smoke== "1", "Yes", "No"))

smoke.1 <- ggplot(cardio_clean_final, aes(x=cardio, fill= smoke.yn)) +
  geom_bar(position="fill") +
  scale_x_discrete(labels=c("1" = "Heart Disease", "0" = "Healthy")) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Tobacco Use", y = "", x = "", fill = "Smoker") +
  scale_fill_manual(values=c("#990000", "#0000da")) +
  theme_minimal()

#alcohol
  cardio_clean_final$alco.yn<- as.factor(ifelse(cardio_clean_final$alco== "1", "Yes", "No"))

alco.1 <- ggplot(cardio_clean_final, aes(x=cardio, fill= alco.yn)) +
  geom_bar(position="fill") +
  scale_x_discrete(labels=c("1" = "Heart Disease", "0" = "Healthy")) +
  scale_y_continuous(labels = scales::percent) +
  labs( title = "Alcohol Use", y = "", x = "", fill = "Drinks Alcohol") +
  scale_fill_manual(values=c("#990000", "#0000da")) +
  theme_minimal()

#active
cardio_clean_final$active.yn<- as.factor(ifelse(cardio_clean_final$active== "1", "Yes", "No"))

active.1 <- ggplot(cardio_clean_final, aes(x=cardio, fill= active.yn)) +
  geom_bar(position="fill") +
  scale_x_discrete(labels=c("1" = "Heart Disease", "0" = "Healthy")) +
  scale_y_continuous(labels = scales::percent) +
  labs( title = "Physical Activity", y = "", x = "", fill = "Physically Active") +
  scale_fill_manual(values=c("#990000", "#0000da")) +
  theme_minimal()

cholesterol.1 + gluc.1 + smoke.1 + gender.1 + alco.1 + active.1 + plot_layout(ncol = 2)
```

**Healthy Patients vs. Heart Disease Patients **  
*According to the plot above, we can draw the conclusion that:*  
- It's close to half the female and male population have cardiovascular disease.  
- The higher the degree of cholesterol and glucose ， more population have this disease.  
- It is surprised that whether patient smokes or not and consumes alcohol or not, Almost half of them have this disease.  
- Do physical exercise can slightly help us decrease the probability of cardiovascular disease.  
- Half of all men and all women have cardiovasulcar disease  
- Higher cholesterol and higher glucose levels can be linked to higher incidence of cardiovascular disease.  
- Almost half of smokers and drinkers have cardiovascular disease.  
- Doing physical exercise helps decrease the probability of getting cardiovascular disease.  

### Statistical Tests
We can use a $\chi^2$ test to determine if the categorical variables and presence of cardiovascular disease are independent.

**Gender**    
- H0: cardio and gender are independent.  
- H1: they are not independent.  

```{r, results='markup'}
cardio_vs_gender = table(cardio_clean_final$cardio, cardio_clean_final$gender)
#rownames(cardio_vs_gender) = c("Absent","Present")
#colnames(cardio_vs_gender) = c("Female", "Male")
#xkabledply(cardio_vs_gender, title="Contingency table for Gender vs Cardiovascular Disease Status")

chi_gender = chisq.test(cardio_vs_gender)
chi_gender
```  
 
Since p-value = `r format(chi_gender$p.value, scientific=T)` > 0.05, we fail to reject the null hypothesis, so `gender` does not have a significant impact on contracting cardiovascular disease.

**Cholesterol**    
- H0: cardio and cholesterol are independent.  
- H1: they are not independent.  
```{r results='markup'}
cardio_vs_cholesterol = table(cardio_clean_final$cardio, cardio_clean_final$cholesterol)
chi_cholesterol = chisq.test(cardio_vs_cholesterol)
chi_cholesterol
```  
* Since p-value = `r format(chi_cholesterol$p.value, scientific=T)` < 0.05, we reject the null hypothesis, so `cholesterol` does have a significant impact on contracting cardiovascular disease.

**Glucose**  
- H0: cardio and gluc are independent.  
- H1: they are not independent.  
```{r results='markup'}
cardio_vs_glucose = table(cardio_clean_final$cardio, cardio_clean_final$gluc)
chi_glucose = chisq.test(cardio_vs_glucose)
chi_glucose
```  
* Since p-value = `r format(chi_glucose$p.value, scientific=T)` < 0.05, we reject the null hypothesis, so `glucose` does have a significant impact on contracting cardiovascular disease.

**Smoking**  
- H0: cardio and smoke are independent.  
- H1: they are not independent.  
```{r results='markup'}
cardio_vs_smoke = table(cardio_clean_final$cardio, cardio_clean_final$smoke)
chi_smoke = chisq.test(cardio_vs_smoke)
chi_smoke
```  
*Since p-value = `r format(chi_smoke$p.value, scientific=T)` < 0.05, we reject the null hypothesis, so `smoking` does have a significant impact on contracting cardiovascular disease.

**Alcohol**  
- H0: cardio and alco are independent.  
- H1: they are not independent.  
```{r results='markup'}
cardio_vs_alco = table(cardio_clean_final$cardio, cardio_clean_final$alco)
chi_alco = chisq.test(cardio_vs_alco)
chi_alco
```  
Since p-value = `r format(chi_alco$p.value, scientific=T)` < 0.05, we reject the null hypothesis, so `alcohol` does have a significant impact on contracting cardiovascular disease.

**Physical Activity**  
- H0: cardio and active are independent.  
- H1: they are not independent.  
```{r results='markup'}
cardio_vs_active = table(cardio_clean_final$cardio, cardio_clean_final$active)
chi_active = chisq.test(cardio_vs_active)
chi_active
```  
Since p-value = `r format(chi_active$p.value, scientific=T)` < 0.05, we reject the null hypothesis, so `activity` does have a significant impact on contracting cardiovascular disease.

**Conclusion:** *Based on the Chi-square tests we performed, we found `gender` does not contribute to the presence or absence of cardiovascular disease. `cholesterol`, `gluc`, `smoke`, `alco` and `active` are the variables that contribute to the presence or absence of cardiovascular disease.*

## Continuous Variables
A brief overview of the dataset tells us we have the following continuous variables: `age`, `height`, `weight`, `ap_hi`, `ap_lo`, plus the new `bmi` variable we added.

### Graphics
The following plots are stacked for better visualization of the distribution. CVD = Cardiovascular Disease.

**Age**
```{r, results='markup'}
cardio_clean_final$cardio.yn<- as.factor(ifelse(cardio_clean_final$active== "1", "Yes", "No"))

age.kde <- ggplot(cardio_clean_final, aes(x = age, fill= cardio.yn)) + 
  geom_density(position = "stack") +
  geom_vline(aes(xintercept=mean(disease$age)), color="steelblue", linetype="dashed", size=1) +
  geom_vline(aes(xintercept=mean(healthy$age)), color="pink3", linetype="dashed", size=1) +
  scale_x_continuous(limits = c(30,70)) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Age Distribution (Stacked)",
       fill = "Has CVD", x = "Age", y = "Distribution") +
  scale_fill_manual(values=c("#990000", "#0000da")) +
  theme_minimal() +
  annotate(geom = "text", x= 56, y =.015, label = "Mean Age of CVD", size = 4,
           hjust = 0, color = "steelblue", fontface = 2) +
  annotate(geom = "text", x= 56, y =.01, label = "Patients is 55", size = 4,
           hjust = 0, color = "steelblue", fontface = 2) +
  annotate(geom = "text", x= 36, y =.09, label = "Mean Age of Healthy", size = 4,
           hjust = 0, color = "pink3", fontface = 2) +
  annotate(geom = "text", x= 36, y =.085, label = "Patients is 52", size = 4,
           hjust = 0, color = "pink3", fontface = 2)
age.kde


```

The average age of CVD patients is 3 years older than the average age of healthy patients. 

**Height**
```{r}
height.kde <- ggplot(cardio_clean_final, aes(x = height, fill= cardio.yn)) + 
  geom_density(position = "stack") +
  geom_vline(aes(xintercept=mean(disease$height)), color="steelblue", linetype="dashed", size=1) +
  geom_vline(aes(xintercept=mean(healthy$height)), color="pink3", linetype="dashed", size=1) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Height Distribution (Stacked)",
       fill = "Has CVD", x = "Height (cm)", y = "Distribution") +
  scale_fill_manual(values=c("#990000", "#0000da")) +
  theme_minimal()

height.kde
```

The average height does not differ between healthy patients and patients with CVD. The average height is 165cm for both groups.

**Weight**
```{r}
weight.kde <- ggplot(cardio_clean_final, aes(x = weight, fill= cardio.yn)) + 
  geom_density(position = "stack") +
  geom_vline(aes(xintercept=mean(disease$weight)), color="steelblue", linetype="dashed", size=1) +
  geom_vline(aes(xintercept=mean(healthy$weight)), color="pink3", linetype="dashed", size=1) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Weight Distribution (Stacked)",
       fill = "Has CVD", x = "Weight (kg)", y = "Distribution") +
  scale_fill_manual(values=c("#990000", "#0000da")) +
  theme_minimal() +
  annotate(geom = "text", x= 99, y =.02, label = "Mean Weight of CVD", size = 4,
           hjust = 0, color = "steelblue", fontface = 2) +
  annotate(geom = "text", x= 99, y =.015, label = "Patients is 75kg", size = 4,
           hjust = 0, color = "steelblue", fontface = 2) +
  annotate(geom = "text", x= 30, y =.072, label = "Mean Weight of Healthy", size = 4,
           hjust = 0, color = "pink3", fontface = 2) +
  annotate(geom = "text", x= 30, y =.067, label = "Patients is 71kg", size = 4,
           hjust = 0, color = "pink3", fontface = 2)

weight.kde
```

The average weight of CVD patients is 4 kg heavier than the average weight of healthy patients. 

**Systolic Blood Pressure**
```{r}
ap_hi.kde <- ggplot(cardio_clean_final, aes(x = ap_hi, fill= cardio.yn)) + 
  geom_density(position = "stack") +
  geom_vline(aes(xintercept=mean(disease$ap_hi)), color="steelblue", linetype="dashed", size=1) +
  geom_vline(aes(xintercept=mean(healthy$ap_hi)), color="pink3", linetype="dashed", size=1) +
  scale_y_continuous(labels = scales::percent) +
  scale_x_continuous(breaks = c(seq(90,170,20))) +
  labs(title = "Systolic BP Distribution (Stacked)",
       fill = "Has CVD", x = "Systolic BP", y = "Distribution") +
  scale_fill_manual(values=c("#990000", "#0000da")) +
  theme_minimal() +
  annotate(geom = "text", x= 135, y =.11, label = "Mean SBP of CVD", size = 4,
           hjust = 0, color = "steelblue", fontface = 2) +
  annotate(geom = "text", x= 135, y =.1, label = "Patients is 133", size = 4,
           hjust = 0, color = "steelblue", fontface = 2) +
  annotate(geom = "text", x= 93, y =.2, label = "Mean SBP of Healthy", size = 4,
           hjust = 0, color = "pink3", fontface = 2) +
  annotate(geom = "text", x= 93, y =.19, label = "Patients is 120", size = 4,
           hjust = 0, color = "pink3", fontface = 2)

ap_hi.kde
```

The average systolic blood pressure (SBP) in CVD patients is 13 units greater than the average systolic blood pressure in healthy patients.

**Diastolic Blood Pressure**
```{r}
ap_lo.kde <- ggplot(cardio_clean_final, aes(x = ap_lo, fill= cardio.yn)) + 
  geom_density(position = "stack") +
  geom_vline(aes(xintercept=mean(disease$ap_lo)), color="steelblue", linetype="dashed", size=1) +
  geom_vline(aes(xintercept=mean(healthy$ap_lo)), color="pink3", linetype="dashed", size=1) +
  scale_y_continuous(labels = scales::percent) +
  scale_x_continuous(breaks = c(seq(65,105,10))) +
  labs(title = "Diastolic BP Distribution (Stacked)",
       fill = "Has CVD", x = "Diastolic BP", y = "Distribution") +
  scale_fill_manual(values=c("#990000", "#0000da")) +
  theme_minimal() +
  annotate(geom = "text", x= 86, y =.25, label = "Mean DBP of CVD", size = 4,
           hjust = 0, color = "steelblue", fontface = 2) +
  annotate(geom = "text", x= 86, y =.22, label = "Patients is 84", size = 4,
           hjust = 0, color = "steelblue", fontface = 2) +
  annotate(geom = "text", x= 66, y =.45, label = "Mean DBP of Healthy", size = 4,
           hjust = 0, color = "pink3", fontface = 2) +
  annotate(geom = "text", x= 66, y =.42, label = "Patients is 79", size = 4,
           hjust = 0, color = "pink3", fontface = 2)

ap_lo.kde
```

The average diastolic blood pressure (DBP) in CVD patients is 5 units greater than the average diastolic blood pressure in healthy patients.

**BMI**
```{r}
bmi.kde <- ggplot(cardio_clean_final, aes(x = bmi, fill= cardio.yn)) + 
  geom_density(position = "stack") +
  geom_vline(aes(xintercept=mean(disease$bmi)), color="steelblue", linetype="dashed", size=1) +
  geom_vline(aes(xintercept=mean(healthy$bmi)), color="pink3", linetype="dashed", size=1) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "BMI Distribution (Stacked)",
       fill = "Has CVD", x = "BMI", y = "Distribution") +
  scale_fill_manual(values=c("#990000", "#0000da")) +
  theme_minimal() +
  annotate(geom = "text", x= 30, y =.13, label = "Mean BMI of CVD", size = 4,
           hjust = 0, color = "steelblue", fontface = 2) +
  annotate(geom = "text", x= 30, y =.12, label = "Patients is 28", size = 4,
           hjust = 0, color = "steelblue", fontface = 2) +
  annotate(geom = "text", x= 15.5, y =.2, label = "Mean BMI of Healthy", size = 4,
           hjust = 0, color = "pink3", fontface = 2) +
  annotate(geom = "text", x= 15.5, y =.19, label = "Patients is 26", size = 4,
           hjust = 0, color = "pink3", fontface = 2)

bmi.kde
```

The average body max index (BMI) of CVD patients is 2 units higher than the average BMI of healthy patients.


###Statistical Tests  
We can use a two-sample t-test to compare if the continuous variable means with and without cardiovascular disease are different.

**Age**  
- H0: cardio_pos and cardio_neg average ages are independent.    
- H1: they are not independent.  
```{r, results='markup'}
cardio_pos = subset(cardio_clean_final, cardio_clean_final$cardio == 1)
cardio_neg = subset(cardio_clean_final, cardio_clean_final$cardio == 0)

ttest_age = t.test(cardio_pos$age, cardio_neg$age)
ttest_age
```  
* Since p-value = `r format(ttest_age$p.value)` < 0.05, we reject the null hypothesis and accept the alternative that the mean `age` of patients with cardiovascular disease is different from patients without cardiovascular disease.
* The p-value is very low

**Height**  
- H0: cardio_pos and cardio_neg average heights are independent.   
- H1: they are not independent.  

```{r, results='markup'}
ttest_height = t.test(cardio_pos$height, cardio_neg$height)
ttest_height
```  
* Since p-value = `r format(ttest_height$p.value)` < 0.05, we reject the null hypothesis and accept the alternative that the mean `height` of patients with cardiovascular disease is different from patients without cardiovascular disease.  However, both of the estimated means are 165cm, and the true difference interval at $a=0.05$ is only -0.364, -0.120.

**Weight**  
- H0: cardio_pos and cardio_neg average weights are independent.  
- H1: they are not independent. 

```{r, results='markup'}
ttest_weight = t.test(cardio_pos$weight, cardio_neg$weight)
ttest_weight
```  
* Since p-value = `r format(ttest_weight$p.value)` < 0.05, we reject the null hypothesis and accept the alternative that the mean `weight` of patients with cardiovascular disease is different from patients without cardiovascular disease.  
* The p-value is very low  

**Systolic Blood Pressure**  
- H0: cardio_pos and cardio_neg average systolic blood pressure are independent.  
- H1: they are not independent. 

```{r, results='markup'}
ttest_ap_hi = t.test(cardio_pos$ap_hi, cardio_neg$ap_hi)
ttest_ap_hi
```
* Since p-value = `r format(ttest_ap_hi$p.value)` < 0.05, we reject the null hypothesis and accept the alternative that the mean `ap_hi` of patients with cardiovascular disease is different from patients without cardiovascular disease.  
* The p-value is very low 

**Diastolic Blood Pressure**  
- H0: cardio_pos and cardio_neg average diastolic blood pressure are independent.    
- H1: they are not independent. 

```{r, results='markup'}
ttest_ap_lo = t.test(cardio_pos$ap_lo, cardio_neg$ap_lo)
ttest_ap_lo
```  
* Since p-value = `r format(ttest_ap_lo$p.value)` < 0.05, we reject the null hypothesis and accept the alternative that the mean `ap_lo` of patients with cardiovascular disease is different from patients without cardiovascular disease.  
* The p-value is very low 

**BMI**  
- H0: cardio_pos and cardio_neg average BMI are independent.   
- H1: they are not independent. 
```{r, results='markup'}
ttest_bmi = t.test(cardio_pos$bmi, cardio_neg$bmi)
ttest_bmi
```  
* Since p-value = `r format(ttest_bmi$p.value)` < 0.05, we reject the null hypothesis and accept the alternative that the mean `bmi` of patients with cardiovascular disease is different from patients without cardiovascular disease.  
* The p-value is very low