---
title: "Team 2 EDA"
author: "Karina Martinez"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r init, include=FALSE}
# some of common options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
library(ezids)
# knitr::opts_chunk$set(warning = F, results = "markup", message = F)
knitr::opts_chunk$set(warning = F, results = "hide", message = F)
options(scientific=T, digits = 3) 
# options(scipen=9, digits = 3) 
# ‘scipen’: integer. A penalty to be applied when deciding to print numeric values in fixed or exponential notation.  Positive values bias towards fixed and negative towards scientific notation: fixed notation will be preferred unless it is more than ‘scipen’ digits wider.
# use scipen=999 to prevent scientific notation at all times
```

# Team 2 Final Project - EDA
This Cardiovascular Disease dataset is from [Kaggle](https://www.kaggle.com/datasets/sulianova/cardiovascular-disease-dataset). The variables are:  

Variable |  Definition  
  :-     |  :---- 
id	 | ID number
age	 | Age in days
gender | Gender (1: women, 2: men)
height | Height in centimeters (cm)
weight | Weight in kilograms (kg)
ap_hi | Systolic blood pressure
ap_lo | Diastolic blood pressure
cholesterol | Cholesterol (1: normal, 2: above normal, 3: well above normal)
gluc | Glucose (1: normal, 2: above normal, 3: well above normal)
smoke | Whether patient smokes or not (0: no, 1: yes)
alco | Whether patient consumes alcohol or not (0: no, 1: yes)
active | Whether patient is physically active or not (0: no, 1: yes)
cardio | Presence or absence of cardiovascular disease (0: absent, 1: present)

```{r, results='markup'}
cardio = data.frame(read.csv("cardio_train.csv", sep=';'))
xkablesummary(cardio, title = "Cardio Data Frame Summary" )
str(cardio)
```  
There are `r nrow(cardio)` observations in the dataset.  
 
Remove the `id` column . The new data frame is saved as `cardio_final`.  
```{r, results='markup'}
#names = c('gender','cholesterol','gluc','smoke','alco','active','cardio')
#cardio_final[,names] <- lapply(cardio[,names] , factor)
cardio_final <- cardio[-c(1)]
#str(cardio_final)
```  

We see that there are unexpected minimum and maximum observations for `height`, `weight`, `ap_hi` and `ap_lo`.  Let's look into these variables.  

```{r}
library(ggplot2)
library(gridExtra)

hist_height = ggplot(cardio_final, aes(height)) + 
  geom_histogram(col="#f67e7d", 
                 fill="#ffb997", 
                 alpha = .9) +
  labs(x="Height (cm)", y="Frequency") +
  labs(title="Height Histogram") 

hist_weight = ggplot(cardio_final, aes(weight)) + 
  geom_histogram(col="blue", 
                 fill="purple", 
                 alpha = .4) + # opacity
  labs(x="Weight (kg)", y="Frequency") +
  labs(title="Weight Histogram") 

hist_hi = ggplot(cardio_final, aes(ap_hi)) + 
  geom_histogram(col="#f67e7d", 
                 fill="#ffb997", 
                 alpha = .9) +
  labs(x="Systolic blood pressure", y="Frequency") +
  labs(title="Systolic blood pressure Histogram") 

hist_lo = ggplot(cardio_final, aes(ap_lo)) + 
  geom_histogram(col="blue", 
                 fill="purple", 
                 alpha = .4) + # opacity
  labs(x="Diastolic blood pressure", y="Frequency") +
  labs(title="Diastolic blood pressure Histogram")

grid.arrange(hist_height, hist_weight, hist_hi, hist_lo, ncol=2, nrow=2)
```  

```{r}
qqnorm(cardio_final$height, main="Q-Q Plot of Height")
qqline(cardio_final$height)

qqnorm(cardio_final$weight, main="Q-Q Plot of Weight")
qqline(cardio_final$weight)

qqnorm(cardio_final$ap_hi, main="Q-Q Plot of Systolic blood pressure")
qqline(cardio_final$ap_hi)

qqnorm(cardio_final$ap_lo, main="Q-Q Plot of Diastolic blood pressure")
qqline(cardio_final$ap_lo)
```  

## Remove outliers

There are outliers above and below the medians for `height` and `weight`.  This is also the case for `ap_hi` and `ap_lo`, but the extremely high outliers are clearly impacting the distribution of these variables. We use the `outlierKD2()` function to remove the outliers.

**Height and Weight**

```{r}
cardio_clean = outlierKD2(cardio_final, height, TRUE, TRUE, TRUE, TRUE) 
cardio_clean = outlierKD2(cardio_clean, weight, TRUE, TRUE, TRUE, TRUE)
```  

**Systolic and Diastolic blood pressure**
```{r}
cardio_clean = outlierKD2(cardio_clean, ap_hi, TRUE, TRUE, TRUE, TRUE) 
cardio_clean = outlierKD2(cardio_clean, ap_lo, TRUE, TRUE, TRUE, TRUE)
``` 

After the outliers were dropped, the QQ-plots for height, weight, and blood pressure improved, indicating the distribution is closer to normal. The "stair-case" pattern on the blood pressure QQ-plots is due to discrete or rounded values. We can also see the more normal distribution in the histograms and box plots for all four variables.  

```{r, results='markup'}
xkablesummary(cardio_clean, title = "Clean Cardio Data Frame Summary" )
cardio_clean = na.omit(cardio_clean) #clean dataframe with numeric variable

#change variables to factor
names <- c('gender','cholesterol','gluc','smoke','alco','active','cardio')
cardio_clean_final <- cardio_clean
cardio_clean_final[,names] <- lapply(cardio_clean[,names] , factor)
```  

We remove the NA rows with the `na.omit()` function and save over the `cardio_clean` data frame. We changed the following variables to categorical (factors): `gender`, `cholesterol`, `gluc`, `smoke`, `alco`, `active` and `cardio` and saved the data frame as `cardio_clean_final`. There are now `r nrow(cardio_clean_final)` observations in the clean data frame.

**Scatterplots**

```{r}
ggplot(subset(cardio_clean_final, cardio_clean_final$gender == 1), aes(x=weight, y=height, col=cardio)) + 
  geom_point() + 
  labs(title="Scatter Plot of Height vs Weight in Women",x="Weight (kg)", y = "Height (cm)", col="Cardiovascular Disease Status")

ggplot(subset(cardio_clean_final, cardio_clean_final$gender == 2), aes(x=weight, y=height, col=cardio)) + 
  geom_point() + 
  labs(title="Scatter Plot of Height vs Weight in Men",x="Weight (kg)", y = "Height (cm)", col="Cardiovascular Disease Status")

ggplot(subset(cardio_clean_final, cardio_clean_final$gender == 1), aes(x=ap_lo, y=ap_hi, col=cardio)) + 
  geom_point() + 
  labs(title="Scatter Plot of Systolic vs Diastolic Blood Pressure in Women",x="Diastolic blood pressure", y = "Systolic blood pressure", col="Cardiovascular Disease Status")

ggplot(subset(cardio_clean_final, cardio_clean_final$gender == 2), aes(x=ap_lo, y=ap_hi, col=cardio)) + 
  geom_point() + 
  labs(title="Scatter Plot of Systolic vs Diastolic Blood Pressure in Men",x="Diastolic blood pressure", y = "Systolic blood pressure", col="Cardiovascular Disease Status")
```  

## Basic Statistics

```{r, results='markup'}

loadPkg("corrplot")
cardio_cor = cor(cardio_clean) # get the correlation matrix  
corrplot.mixed(cardio_cor) 
```  

**T-tests**

```{r, results='markup'}
cardio_pos = subset(cardio_clean_final, cardio_clean_final$cardio == 1)
cardio_neg = subset(cardio_clean_final, cardio_clean_final$cardio == 0)

ttest_height = t.test(cardio_pos$height, cardio_neg$height)
ttest_height

ttest_weight = t.test(cardio_pos$weight, cardio_neg$weight)
ttest_weight
```
```{r}

box_weight_pos = ggplot(cardio_pos, aes(y=weight)) + 
  geom_boxplot() + 
  geom_boxplot(colour="#f67e7d", fill="#ffb997", alpha = .9, outlier.colour="red", outlier.size=2) +
  labs(title="Weight Boxplot (Cardiovascular \nDisease Present)",x="", y = "Weight (kg)")

box_weight_neg = ggplot(cardio_neg, aes(y=weight)) + 
  geom_boxplot() + 
  geom_boxplot(colour="blue", fill="purple", alpha = .4,  outlier.colour="red", outlier.size=2) +
  labs(title="Weight Boxplot (Cardiovascular \nDisease Absent)",x="", y = "Weight (kg)")

grid.arrange(box_weight_pos, box_weight_neg, ncol=2, nrow=1)
```

```{r, results='markup'}
ttest_hi = t.test(cardio_pos$ap_hi, cardio_neg$ap_hi)
ttest_hi

ttest_lo = t.test(cardio_pos$ap_lo, cardio_neg$ap_lo)
ttest_lo
```
```{r, results='markup'}
anovaRes = aov(weight ~ gluc, data=cardio_clean_final)
xkabledply(anovaRes, title = "ANOVA result summary")
```

```{r, results='markup'}
anovaRes = aov(weight ~ cholesterol, data=cardio_clean_final)
xkabledply(anovaRes, title = "ANOVA result summary")
```

```{r, results='markup'}
contable = table(cardio_clean_final$gender, cardio_clean_final$cardio)
rownames(contable) = c("Female", "Male")
colnames(contable) = c("Absent","Present")

xkabledply(contable, title="Contingency table for Gender vs Cardiovascular Disease Status")
```

```{r, results='markup'}
chitest = chisq.test(contable, correct=FALSE)
chitest
```  
In this data set, men are not more likely than women to have cardiovascular disease.

  
