---
title: "Team 2 Technical Analysis"
author: "Hovhannes Arestakesyan, Karina Martinez, Timberley Thompson, Chenxu Wang, & Zhe Yu"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r setup, include=FALSE}
#Libraries
library(ezids)
library(tidyverse)

#Knit Options
knitr::opts_chunk$set(warning = F, results = "hide", message = F)
options(scientific=T, digits = 3) 
```

# Cardiovascular Dataset
```{r}
#Import Data
cardio_orig = data.frame(read.csv("cardio_train.csv", header = TRUE, sep=';'))
str(cardio_orig)
```
This Cardiovascular Disease dataset is from [Kaggle](https://www.kaggle.com/datasets/sulianova/cardiovascular-disease-dataset). The variables are:  

Variable |  Definition  
  :-     |  :---- 
id	 | ID number
age	 | Age in days
gender | Gender (1: women, 2: men)
height | Height in centimeters (cm)
weight | Weight in kilograms (kg)
ap_hi | Systolic blood pressure
ap_lo | Diastolic blood pressure
cholesterol | Cholesterol (1: normal, 2: above normal, 3: well above normal)
gluc | Glucose (1: normal, 2: above normal, 3: well above normal)
smoke | Whether patient smokes or not (0: no, 1: yes)
alco | Whether patient consumes alcohol or not (0: no, 1: yes)
active | Whether patient is physically active or not (0: no, 1: yes)
cardio | Presence or absence of cardiovascular disease (0: absent, 1: present)

There are 70000 observations in the dataset.

# Data Cleanup
```{r}
#Fix Variables
cardio = cardio_orig[-c(1)] #get rid of id
cardio$age = cardio$age/365 # transfer days to years
cardio$gender = as.factor(cardio$gender)
cardio$cholesterol = as.factor(cardio$cholesterol) 
cardio$gluc = as.factor(cardio$gluc)
cardio$smoke = as.factor(cardio$smoke)
cardio$alco = as.factor(cardio$alco)
cardio$active = as.factor(cardio$active)
cardio$cardio = as.factor(cardio$cardio)
cardio$bmi = cardio$weight/((cardio$height/100)**2) #add bmi variable

summary(cardio)
```

We remove the `id` column and converted `age` from days to years.  We added a new column `bmi`, for body mass index. Body mass index (BMI) is a measure of body fat based on height and weight, and the formula is $BMI = kg/m^2$. We also converted categorical variables from integer to factor. The new data frame is saved as `cardio`. 

We see that there are unexpected minimum and maximum observations for `height`, `weight`, `ap_hi`, `ap_lo` and `bmi`.  Blood pressure measurements as low as -150 and as high as 16020 are incompatable with life, as are BMIs of 3.5 and 299.  We use the `outlierKD2()` and `na.omit()` functions to remove the observations with outliers. 

```{r}
#Remove Outliers
cardio_clean = outlierKD2(cardio, bmi, TRUE, TRUE, TRUE, TRUE) 
cardio_clean = outlierKD2(cardio_clean, ap_hi, TRUE, TRUE, TRUE, TRUE) 
cardio_clean = outlierKD2(cardio_clean, ap_lo, TRUE, TRUE, TRUE, TRUE)

cardio_clean_final = na.omit(cardio_clean) #remove NA rows
```

After the outliers were dropped, the QQ-plots improved, indicating the distribution is closer to normal. The "stair-case" pattern on the blood pressure QQ-plots is due to discrete or rounded values. We can also see the more normal distribution in the histograms and box plots for all of the variables.  
The new dataframe is saved as `cardio_clean_final`. There are `r nrow(cardio_clean_final)` observations in the new dataset.

```{r}
xkablesummary(cardio_clean_final, title = "Clean Cardio Data Frame Summary" )
```

# Exploratory Data Analysis (EDA)

```{r}
disease <- subset(cardio_clean_final, cardio==1)
healthy <- subset(cardio_clean_final, cardio==0)
summary(disease)
summary(healthy)

```

## Categorical Variables
### Graphics
```{r, results='markup'}
require(patchwork)

#cholesterol
cholesterol.1 <- ggplot(cardio_clean_final, aes(x=cardio, fill= cholesterol)) +
  geom_bar(position="fill") +
  scale_x_discrete(labels=c("1" = "Heart Disease", "0" = "Healthy")) +
  scale_y_continuous(labels = scales::percent) +
  labs( title = "Cholesterol", y = "", x = "",
        fill = "Cholesterol Level") +
  scale_fill_manual(values=c("#f4cccc", "#e06666", "#990000")) +
  theme_minimal()

#Glucose
gluc.1 <- ggplot(cardio_clean_final, aes(x=cardio, fill= gluc)) +
  geom_bar(position="fill") +
  scale_x_discrete(labels=c("1" = "Heart Disease", "0" = "Healthy")) +
  scale_y_continuous(labels = scales::percent) +
  labs( title = "Glucose", y = "", x = "",
        fill = "Glucose Level") +
  scale_fill_manual(values=c("#f4cccc", "#e06666", "#990000")) +
  theme_minimal()

#Gender
cardio_clean_final$gender.yn<- as.factor(ifelse(cardio_clean_final$gender== "1", "Women", "Men"))

gender.1 <- ggplot(cardio_clean_final, aes(x=cardio, fill= gender.yn)) +
  geom_bar(position="fill") +
  scale_x_discrete(labels=c("1" = "Heart Disease", "0" = "Healthy")) +
  scale_y_continuous(labels = scales::percent) +
  labs( title = "Gender",
        y = "", x = "", fill = "") +
  scale_fill_manual(values=c("#990000", "#0000da")) +
  theme_minimal()

#Smoke
cardio_clean_final$smoke.yn<- as.factor(ifelse(cardio_clean_final$smoke== "1", "Yes", "No"))

smoke.1 <- ggplot(cardio_clean_final, aes(x=cardio, fill= smoke.yn)) +
  geom_bar(position="fill") +
  scale_x_discrete(labels=c("1" = "Heart Disease", "0" = "Healthy")) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Tobacco Use", y = "", x = "", fill = "Smoker") +
  scale_fill_manual(values=c("#990000", "#0000da")) +
  theme_minimal()

#alcohol
  cardio_clean_final$alco.yn<- as.factor(ifelse(cardio_clean_final$alco== "1", "Yes", "No"))

alco.1 <- ggplot(cardio_clean_final, aes(x=cardio, fill= alco.yn)) +
  geom_bar(position="fill") +
  scale_x_discrete(labels=c("1" = "Heart Disease", "0" = "Healthy")) +
  scale_y_continuous(labels = scales::percent) +
  labs( title = "Alcohol Use", y = "", x = "", fill = "Drinks Alcohol") +
  scale_fill_manual(values=c("#990000", "#0000da")) +
  theme_minimal()

#active
cardio_clean_final$active.yn<- as.factor(ifelse(cardio_clean_final$active== "1", "Yes", "No"))

active.1 <- ggplot(cardio_clean_final, aes(x=cardio, fill= active.yn)) +
  geom_bar(position="fill") +
  scale_x_discrete(labels=c("1" = "Heart Disease", "0" = "Healthy")) +
  scale_y_continuous(labels = scales::percent) +
  labs( title = "Physical Activity", y = "", x = "", fill = "Physically Active") +
  scale_fill_manual(values=c("#990000", "#0000da")) +
  theme_minimal()

cholesterol.1 + gluc.1 + smoke.1 + gender.1 + alco.1 + active.1 + plot_layout(ncol = 2)
```

**Healthy Patients vs. Heart Disease Patients **  
*According to the plot above, we can draw the conclusion that: *  
- It's close to half the female and male population have cardiovascular disease.  
- The higher the degree of cholesterol and glucose ， more population have this disease.  
- It is surprised that whether patient smokes or not and consumes alcohol or not, Almost half of them have this disease.  
- Do physical exercise can slightly help us decrease the probability of cardiovascular disease.  
- Half of all men and all women have cardiovasulcar disease  
- Higher cholesterol and higher glucose levels can be linked to higher incidence of cardiovascular disease.  
- Almost half of smokers and drinkers have cardiovascular disease.  
- Doing physical exercise helps decrease the probability of getting cardiovascular disease.  

### Statistical Tests
We can use a $\chi^2$ test to determine if the categorical variables and presence of cardiovascular disease are independent.

**Gender**
**1. Whether `cardio` and `gender` are independent.**  
- H0: cardio and gender are independent.  
- H1: they are not independent.  

```{r, results='markup'}
cardio_vs_gender = table(cardio_clean_final$cardio, cardio_clean_final$gender)
#rownames(cardio_vs_gender) = c("Absent","Present")
#colnames(cardio_vs_gender) = c("Female", "Male")
#xkabledply(cardio_vs_gender, title="Contingency table for Gender vs Cardiovascular Disease Status")

chi_gender = chisq.test(cardio_vs_gender)
chi_gender
```  
* In this data set, there gender and cardiovascular disease are NOT significantly independent. 
* The p-value from the Chi-square test is 0.2, which is larger than 0.05. So we failed to reject the null hypothesis. This result indicating `gender` difference won't affect the onset of  cardiovascular diseases. 
* Since p-value = `r format(chi_test1$p.value, scientific=T)` > 0.05, we fail to reject the null hypothesis, so `gender` does not have a significant impact on contracting cardiovascular disease.

### Cholesterol
**2. Whether `cardio` and `cholesterol` are independent.**  
- H0: cardio and cholesterol are independent.  
- H1: they are not independent.  
```{r results='markup'}
cardio_vs_cholesterol = table(cardio_clean_final$cardio, cardio_clean_final$cholesterol)
chi_cholesterol = chisq.test(cardio_vs_cholesterol)
chi_cholesterol
```  
* The p-value from the Chi-square test is less than 2e-16, which is smaller than 0.05. So we can reject the null hypothesis. This result indicating `cholesterol` level will contribute to the onset of cardiovascular diseases.  
* Since p-value = `r format(chi_test1$p.value, scientific=T)` < 0.05, we reject the null hypothesis, so `cholesterol` does have a significant impact on contracting cardiovascular disease.

### Glucose
**3. Whether `cardio` and `gluc` are independent.**  
- H0: cardio and gluc are independent.  
- H1: they are not independent.  
```{r results='markup'}
cardio_vs_glucose = table(cardio_clean_final$cardio, cardio_clean_final$gluc)
chi_glucose = chisq.test(cardio_vs_glucose)
chi_glucose
```  
* The p-value from the Chi-square test is less than 2e-16, which is smaller than 0.05. So we can reject the null hypothesis. This result indicating glucose (`gluc`) level will contribute to the onset of cardiovascular diseases. 
* Since p-value = `r format(chi_test1$p.value, scientific=T)` < 0.05, we reject the null hypothesis, so `glucose` does have a significant impact on contracting cardiovascular disease.

### Smoking
**4. Whether `cardio` and `smoke` are independent.**  
- H0: cardio and smoke are independent.  
- H1: they are not independent.  
```{r results='markup'}
cardio_vs_smoke = table(cardio_clean_final$cardio, cardio_clean_final$smoke)
chi_smoke = chisq.test(cardio_vs_smoke)
chi_smoke
```  
*The p-value from the Chi-square test is 3e-08, which is smaller than 0.05. So we can reject the null hypothesis. This result indicating smoke or not will contribute to the onset of cardiovascular diseases.  
*Since p-value = `r format(chi_test1$p.value, scientific=T)` < 0.05, we reject the null hypothesis, so `smoking` does have a significant impact on contracting cardiovascular disease.

### Alcohol
**5. Whether `cardio` and `alco` are independent.**  
- H0: cardio and alco are independent.  
- H1: they are not independent.  
```{r results='markup'}
cardio_vs_alco = table(cardio_clean_final$cardio, cardio_clean_final$alco)
chi_alco = chisq.test(cardio_vs_alco)
chi_alco
```  
* The p-value from the Chi-square test is 0.002, which is smaller than 0.05. So we can reject the null hypothesis. This result indicating whether the patient consumes alcohol or not will contribute to the onset of cardiovascular diseases. 
* Since p-value = `r format(chi_test1$p.value, scientific=T)` < 0.05, we reject the null hypothesis, so `alcohol` does have a significant impact on contracting cardiovascular disease.

### Physical Activity
**6. Whether `cardio` and `active` are independent.**  
- H0: cardio and active are independent.  
- H1: they are not independent.  
```{r results='markup'}
cardio_vs_active = table(cardio_clean_final$cardio, cardio_clean_final$active)
chi_active = chisq.test(cardio_vs_active)
chi_active
```  
* The p-value from the Chi-square test is less then 2e-16, which is smaller than 0.05. So we can reject the null hypothesis. This result indicating whether the patient is physically active or not will contribute to the onset of cardiovascular diseases.  
* Since p-value = `r format(chi_test1$p.value, scientific=T)` < 0.05, we reject the null hypothesis, so `activity` does have a significant impact on contracting cardiovascular disease.

**Based on the Chi-square tests we performed, we found `gender` is not contribute to the presence or absence of cardiovascular disease, while `cholesterol`, `gluc`, `smoke`, `alco` and `active` are the variables that can affect whether the patient has cardiovascular disease or not.**



